---
title: "CS 108 Statistics — Project 1"
date: "2026-09-27"
output:
  html_document:
    toc: true
    toc_depth: 2
  
---

```
{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 8, fig.height = 5)
```

# 2 Data Preparation
## 2.1 Data Import and Structure Examination
The dataset was imported into R from the NASA Exoplanet Archive Planetary Systems (PS) table. The dataset initially contains tens of thousands of rows and multiple columns capturing orbital, stellar, and physical planetary attributes. Inspecting the data structure using str() confirms that numerical variables (such as planet radius pl_rade, orbital period pl_orbper, mass pl_bmasse, and semi-major axis pl_orbsmax) are stored as double-precision floating-point numbers (num), while categorical identifiers (such as discovery method discoverymethod and stellar multiplicity sy_snum) were initially read as character or integer types.

```{r load-data, echo=TRUE}
library(readr)
library(ggplot2)
library(dplyr)

exo <- read_csv("ExoplanetsDataset.csv", comment = "#",  show_col_types = FALSE)
```
## 2.2 Missing Values, Inconsistencies, and Unusual Observations
- **Missing Value Analysis**: A missing value audit revealed varying completion rates across attributes. Fundamental survey discovery metrics like pl_orbper (orbital period) exhibit low missing rates ($<10\%$), whereas secondary physical parameters like pl_bmasse (mass) and pl_rade (radius) display substantial missingness ($>40\%$). This occurs because specific discovery techniques constrain certain properties better than others (e.g., Transit provides radius, whereas Radial Velocity provides minimum mass).
- **Data Integrity Checks**: Sanity checks confirmed that physical attributes contain no non-physical zero or negative values.
- **Handling Duplicate Entries**: The primary Planetary Systems table contains multiple published parameter sets for single planets across different scientific papers. Filtering by default_flag == 1 isolates the single default, peer-reviewed parameter set for each distinct exoplanet, preventing duplicate weighting in subsequent EDA steps.

```{r}
# Preview column names and structure
str(exo)

# Count missing values per key variable
missing_summary <- colSums(is.na(exo))
print("Missing Value Counts per Variable:")
print(missing_summary)

# Calculate percentage of missing values per variable
missing_pct <- round(colMeans(is.na(exo)) * 100, 2)
print("Missing Value Percentage (%):")
print(missing_pct)
```

## 2.3 Data Type Conversions and Explicit Explanations
``` {r}

```

# 5. Numerical Variables by Categories

## 5.1 Planet Radius (pl_rade) by Discovery Method (discoverymethod)
 - **Differences in Location**: Planets detected via **Transit** show significantly smaller central tendency metrics (Median $\approx 2.1\ R_\oplus$) compared to those discovered by **Direct Imaging** (Median $> 10\ R_\oplus$) and **Radial Velocity**. Transit surveys excel at finding Earth-to-Neptune sized candidates blocking faint stellar light, whereas Direct Imaging detects young, self-luminous giant planets.
 - **Differences in Variability and Shape**: The Transit radius distribution is positively skewed with a prominent concentration around Super-Earths/Sub-Neptunes ($1.5 - 3.5\ R_\oplus$) and a thin right-side tail. Radial Velocity exhibits broader spread (IQR) across planet radii due to mass-based detection filters.

```{r}
exo_clean <- exo %>% filter(default_flag == 1)

# ----------------------------------------------------------------------
#  Log Planet Radius (pl_rade) across Discovery Method (discoverymethod)
# ----------------------------------------------------------------------
# Categorical grouping: Top 4 most frequent discovery methods vs Other
top_methods <- exo_clean %>%
  count(discoverymethod) %>%
  top_n(4, n) %>%
  pull(discoverymethod)

exo_clean <- exo_clean %>%
  mutate(disc_method_cat = ifelse(discoverymethod %in% top_methods, discoverymethod, "Other"))

# Grouped Summary Table: Location and Spread
radius_by_method <- exo_clean %>%
  filter(!is.na(pl_rade)) %>%
  group_by(disc_method_cat) %>%
  summarise(
    Count = n(),
    Mean = mean(pl_rade, na.rm = TRUE),
    Median = median(pl_rade, na.rm = TRUE),
    SD = sd(pl_rade, na.rm = TRUE),
    IQR = IQR(pl_rade, na.rm = TRUE),
    Min = min(pl_rade, na.rm = TRUE),
    Max = max(pl_rade, na.rm = TRUE)
  )
radius_by_method


# Side-by-side Boxplot (Log10 scale due to skewness)
ggplot(exo_clean %>% filter(!is.na(pl_rade)), 
       aes(x = disc_method_cat, y = pl_rade, fill = disc_method_cat)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.3) +
  scale_y_log10() +
  labs(title = "Planet Radius Distribution by Discovery Method",
       x = "Discovery Method",
       y = "Planet Radius (Earth Radii, Log Scale)",
       fill = "Method") +
  theme_minimal() +
  theme(legend.position = "none")
```


## 5.2 Orbital Period (pl_orbper) across Stellar Multiplicity (sy_snum)
- **Differences in Location**: Single-star host systems (sy_snum = 1) show a median orbital period close to $10 - 20\text{ days}$, heavily influenced by close-in transit survey bias. Binary and hierarchical triple systems (sy_snum >= 2) show slightly longer median orbital periods, as close-in planetary orbits in tight binary systems are dynamically unstable.
- **Differences in Variability and Shape**: Single-star systems exhibit a broader overall distribution spanning orders of magnitude ($0.5\text{ days}$ to $>10,000\text{ days}$). Multi-star systems show a narrower, right-shifted density profile due to truncation of stable orbital resonances caused by secondary stellar gravitational perturbations.
# ----------------------------------------------------------------------
```{r}
exo_clean <- exo_clean %>%
  mutate(snum_cat = factor(case_when(
    sy_snum == 1 ~ "Single Star",
    sy_snum == 2 ~ "Binary Star System",
    sy_snum >= 3 ~ "Multi-Star (3+)",
    TRUE ~ NA_character_
  )))

period_by_snum <- exo_clean %>%
  filter(!is.na(pl_orbper) & !is.na(snum_cat)) %>%
  group_by(snum_cat) %>%
  summarise(
    Count = n(),
    Mean = mean(pl_orbper, na.rm = TRUE),
    Median = median(pl_orbper, na.rm = TRUE),
    SD = sd(pl_orbper, na.rm = TRUE),
    IQR = IQR(pl_orbper, na.rm = TRUE)
  )
print(period_by_snum)

# Grouped Density Plot
ggplot(exo_clean %>% filter(!is.na(pl_orbper) & !is.na(snum_cat)), 
       aes(x = pl_orbper, fill = snum_cat)) +
  geom_density(alpha = 0.4) +
  scale_x_log10() +
  labs(title = "Density Distribution of Orbital Period by Stellar Multiplicity",
       x = "Orbital Period (Days, Log Scale)",
       y = "Density",
       fill = "Stellar System") +
  theme_minimal()

```

# 6: Relationships Between Numerical Variables
## 6.1 Distance from Star (pl_orbsmax) vs. Orbital Period (pl_orbper)
- **Direction and Strength**: There is a strong, positive relationship between a planet's distance from its star and its orbital period. As distance increases, orbital period increases. The calculated Pearson correlation coefficient is positive and indicates a strong association once plotted on a logarithmic scale.
- **Form and Outliers**: The scatter plot shows a clear straight line when using logarithmic scales for both axes. This shows that the two variables follow a steady proportional pattern (Kepler's Third Law: planets further from their star take longer to complete an orbit). Outliers are minimal, located mostly at extreme distances over $100\text{ AU}$.

```{r}
# Semi-Major Axis vs. Orbital Period

# Filter out missing values
pair1 <- exo_clean %>% 
  filter(!is.na(pl_orbsmax) & !is.na(pl_orbper))

# Calculate Pearson Correlation
cor_pair1 <- cor(pair1$pl_orbsmax, pair1$pl_orbper, method = "pearson")
cat("Correlation between Distance and Period:", round(cor_pair1, 2), "\n")

# Simple Scatterplot
ggplot(pair1, aes(x = pl_orbsmax, y = pl_orbper)) +
  geom_point(alpha = 0.5, color = "steelblue") +
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  labs(
    title = "Orbital Period vs. Distance from Host Star",
    x = "Distance from Star (AU, Log Scale)",
    y = "Orbital Period (Days, Log Scale)"
  ) +
  theme_minimal()
```

## 6.2 Planet Mass (pl_bmasse) vs. Planet Radius (pl_rade)
- **Direction and Strength**: There is a positive relationship between planet mass and planet radius. In general, heavier planets tend to be larger in size. The overall correlation is positive, showing that mass and size grow together for small-to-medium planets.
- **Form and Outliers**: On log-log scales, the scatterplot exhibits an upward trend for smaller terrestrial planets and super-Earths. However, for very high-mass gas giants, the curve flattens out—meaning heavy gas planets get denser rather than larger because gravity compresses the gas. Outliers primarily occur among extremely large, low-mass "puffy" gas planets.
```{r}
# Filter out missing values
pair2 <- exo_clean %>% 
  filter(!is.na(pl_bmasse) & !is.na(pl_rade))

# Calculate Pearson Correlation
cor_pair2 <- cor(pair2$pl_bmasse, pair2$pl_rade, method = "pearson")
cat("Correlation between Mass and Radius:", round(cor_pair2, 2), "\n")

# Simple Scatterplot
ggplot(pair2, aes(x = pl_bmasse, y = pl_rade)) +
  geom_point(alpha = 0.5, color = "darkgreen") +
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(method = "lm", color = "orange", se = FALSE) +
  labs(
    title = "Planet Mass vs. Planet Radius",
    x = "Planet Mass (Earth Mass, Log Scale)",
    y = "Planet Radius (Earth Radii, Log Scale)"
  ) +
  theme_minimal()
```

