---
title: "CS 108 Statistics — Project 1"
date: "2026-09-27"
output:
  html_document:
    toc: true
    toc_depth: 2
  
---

```
{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 8, fig.height = 5)
```

# 2 Data Preparation
## 2.1 Data Import and Structure Examination
The dataset was imported into R from the NASA Exoplanet Archive Planetary Systems (PS) table. The dataset initially contains tens of thousands of rows and multiple columns capturing orbital, stellar, and physical planetary attributes. Inspecting the data structure using str() confirms that numerical variables (such as planet radius pl_rade, orbital period pl_orbper, mass pl_bmasse, and semi-major axis pl_orbsmax) are stored as double-precision floating-point numbers (num), while categorical identifiers (such as discovery method discoverymethod and stellar multiplicity sy_snum) were initially read as character or integer types.

```{r load-data, echo=TRUE}
library(readr)
library(ggplot2)
library(dplyr)

exo <- read_csv("ExoplanetsDataset.csv", comment = "#",  show_col_types = FALSE)

dim(exo)
names(exo)
str(exo)
```
## 2.2 Missing Values, Inconsistencies, and Unusual Observations
- **Missing Value Analysis**: The missing-value analysis showed that the completeness of variables differs substantially. Orbital period has fewer missing values than several physical properties such as planet mass and radius. Because missingness varies across variables, analyses involving these variables use only observations for which the required variables are available.
```{r}
key_vars <- c(
  "pl_rade",
  "pl_orbper",
  "pl_bmasse",
  "pl_orbsmax",
  "discoverymethod",
  "sy_snum"
)

missing_summary <- data.frame(
  Variable = key_vars,
  Missing = sapply(exo[key_vars], function(x) sum(is.na(x))),
  Missing_Percent = round(
    sapply(exo[key_vars], function(x) mean(is.na(x))) * 100, 2
  )
)

missing_summary
```
- **Data Integrity Checks**: Sanity checks confirmed that physical attributes contain no non-physical zero or negative values.
```{r}
sum(exo$pl_rade <= 0, na.rm = TRUE)
sum(exo$pl_orbper <= 0, na.rm = TRUE)
sum(exo$pl_bmasse <= 0, na.rm = TRUE)
sum(exo$pl_orbsmax <= 0, na.rm = TRUE)
```
- **Handling Duplicate Entries**: The primary Planetary Systems table contains multiple published parameter sets for single planets across different scientific papers. Filtering by default_flag == 1 isolates the single default, peer-reviewed parameter set for each distinct exoplanet, preventing duplicate weighting in subsequent EDA steps.
```{r}
sum(exo$default_flag == 1, na.rm = TRUE)
```


## 2.3 Data Type Conversions and Explicit Explanations
``` {r}

```


# 5. Numerical Variables by Categories

## 5.1 Planet Radius (pl_rade) by Discovery Method (discoverymethod)
 - **Differences in Location**: The median planet radius differs across the major discovery-method groups. The Transit group has a lower median radius than some of the other discovery methods. The median is useful for this comparison because planet radius has a skewed distribution and can contain extreme values.
 - **Differences in Variability and Shape**:The distributions also differ in their spread and shape. The Transit group has many observations concentrated at smaller planet radii, with a longer tail toward larger radii. The boxplot also shows differences in the IQR and the presence of potential outliers across discovery methods.
 To make the comparison readable, the four most frequent discovery methods were retained as separate categories and all remaining methods were combined into an "Other" category. This reduces the number of groups shown in the graph, but the "Other" category contains heterogeneous discovery methods and therefore should not be interpreted as a single detection technique.

```{r}
exo_clean <- exo %>% filter(default_flag == 1)

# ----------------------------------------------------------------------
#  Log Planet Radius (pl_rade) across Discovery Method (discoverymethod)
# ----------------------------------------------------------------------
# Categorical grouping: Top 4 most frequent discovery methods vs Other
top_methods <- exo_clean %>%
  count(discoverymethod) %>%
  top_n(4, n) %>%
  pull(discoverymethod)

exo_clean <- exo_clean %>%
  mutate(disc_method_cat = ifelse(discoverymethod %in% top_methods, discoverymethod, "Other"))

# Grouped Summary Table: Location and Spread
radius_by_method <- exo_clean %>%
  filter(!is.na(pl_rade)) %>%
  group_by(disc_method_cat) %>%
  summarise(
    Count = n(),
    Mean = mean(pl_rade, na.rm = TRUE),
    Median = median(pl_rade, na.rm = TRUE),
    SD = sd(pl_rade, na.rm = TRUE),
    IQR = IQR(pl_rade, na.rm = TRUE),
    Min = min(pl_rade, na.rm = TRUE),
    Max = max(pl_rade, na.rm = TRUE)
  )
radius_by_method


# Side-by-side Boxplot (Log10 scale due to skewness)
ggplot(exo_clean %>% filter(!is.na(pl_rade)), 
       aes(x = disc_method_cat, y = pl_rade, fill = disc_method_cat)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.3) +
  scale_y_log10() +
  labs(title = "Planet Radius Distribution by Discovery Method",
       x = "Discovery Method",
       y = "Planet Radius (Earth Radii, Log Scale)",
       fill = "Method") +
  theme_minimal() +
  theme(legend.position = "none")

Q1 <- quantile(exo_clean$pl_rade, 0.25, na.rm = TRUE)
Q3 <- quantile(exo_clean$pl_rade, 0.75, na.rm = TRUE)
IQR_value <- Q3 - Q1

outliers <- exo_clean %>%
  filter(
    pl_rade < Q1 - 1.5 * IQR_value |
    pl_rade > Q3 + 1.5 * IQR_value
  )

outliers %>%
  select(pl_name, pl_rade, discoverymethod)
```


## 5.2 Orbital Period (pl_orbper) across Stellar Multiplicity (sy_snum)
The distribution of orbital period was compared across planetary systems with different numbers of stars. The sy_snum variable was grouped into three categories: Single Star (sy_snum = 1), Binary Star System (sy_snum = 2), and Multi-Star (3+) (sy_snum ≥ 3). Only observations with available orbital-period values were included in this comparison.
### Differences in Location
The median orbital period is 10.6 days for Single Star systems, 12.8 days for Binary Star Systems, and 13.7 days for Multi-Star systems. Therefore, the typical orbital period is somewhat longer in systems with two or more stars.
The means are much larger than the medians in all three groups. For example, the mean for Binary Star Systems is 16,857 days, compared with a median of only 12.8 days. This large difference suggests that a relatively small number of planets with very long orbital periods strongly affect the mean. Therefore, the median is more useful than the mean for comparing the typical orbital period across these groups.
### Differences in Variability and Shape
There are substantial differences in variability between the groups. The IQR is 30.8 days for Single Star systems, 213 days for Binary Star Systems, and 252 days for Multi-Star systems. This shows that the middle 50% of orbital periods are much more spread out in the Binary and Multi-Star groups.
The standard deviations are also very large, especially for Binary Star Systems. The standard deviation is 147,179 days for Single Star systems, 278,352 days for Binary Star systems, and 12,515 days for Multi-Star systems. These large values indicate that extreme orbital-period observations have a strong effect on the standard deviation.

Overall, Most discovery-method groups have right-skewed distributions, as shown by the large differences between the means and medians. The boxplot and density plot on the logarithmic scale make differences in spread and the presence of extreme values easier to see. These results show an association between stellar multiplicity and the distribution of orbital periods, but they do not establish that the number of stars causes differences in orbital period.
```{r}
exo_clean <- exo_clean %>%
  mutate(snum_cat = factor(case_when(
    sy_snum == 1 ~ "Single Star",
    sy_snum == 2 ~ "Binary Star System",
    sy_snum >= 3 ~ "Multi-Star (3+)",
    TRUE ~ NA_character_
  )))

period_by_snum <- exo_clean %>%
  filter(!is.na(pl_orbper) & !is.na(snum_cat)) %>%
  group_by(snum_cat) %>%
  summarise(
    Count = n(),
    Mean = mean(pl_orbper, na.rm = TRUE),
    Median = median(pl_orbper, na.rm = TRUE),
    SD = sd(pl_orbper, na.rm = TRUE),
    IQR = IQR(pl_orbper, na.rm = TRUE)
  )
print(period_by_snum)

# Grouped Density Plot
ggplot(exo_clean %>% filter(!is.na(pl_orbper) & !is.na(snum_cat)), 
       aes(x = pl_orbper, fill = snum_cat)) +
  geom_density(alpha = 0.4) +
  scale_x_log10() +
  labs(title = "Density Distribution of Orbital Period by Stellar Multiplicity",
       x = "Orbital Period (Days, Log Scale)",
       y = "Density",
       fill = "Stellar System") +
  theme_minimal()

ggplot(
  exo_clean %>% filter(!is.na(pl_orbper), !is.na(snum_cat)),
  aes(x = snum_cat, y = pl_orbper)
) +
  geom_boxplot() +
  scale_y_log10() +
  labs(
    title = "Orbital Period by Stellar Multiplicity",
    x = "Stellar System",
    y = "Orbital Period (Days, Log Scale)"
  ) +
  theme_minimal()

```

# 6: Relationships Between Numerical Variables
## 6.1 Distance from Star (pl_orbsmax) vs. Orbital Period (pl_orbper)
- **Direction and Strength**: There is a strong positive relationship between a planet's semi-major axis and its orbital period. In general, planets with larger semi-major axes tend to have longer orbital periods. The Pearson correlation coefficient indicates a strong positive association between the two variables.
- **Form and Outliers**: On the logarithmic scales, the scatterplot shows a clear positive and approximately linear pattern.This indicates that the relationship is well described by a power-law-type pattern on the log-log scale. A small number of observations appear far from the main concentration of points, particularly at very large semi-major axes.

```{r}
# Semi-Major Axis vs. Orbital Period

# Filter out missing values, i.e complete-case analysis for this pair.
pair1 <- exo_clean %>% 
  filter(!is.na(pl_orbsmax) & !is.na(pl_orbper))

# Calculate Pearson Correlation
cor_pair1 <- cor(pair1$pl_orbsmax, pair1$pl_orbper, method = "pearson")
cat("Correlation between Distance and Period:", round(cor_pair1, 2), "\n")

# Simple Scatterplot
ggplot(pair1, aes(x = pl_orbsmax, y = pl_orbper)) +
  geom_point(alpha = 0.5, color = "steelblue") +
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  labs(
    title = "Orbital Period vs. Distance from Host Star",
    x = "Distance from Star (AU, Log Scale)",
    y = "Orbital Period (Days, Log Scale)"
  ) +
  theme_minimal()
```

## 6.2 Planet Mass (pl_bmasse) vs. Planet Radius (pl_rade)
- **Direction and Strength**: There is a positive relationship between planet mass and planet radius. In general, planets with larger masses tend to have larger radii, although the points show considerable variation around the overall trend.
- **Form and Outliers**: On the log-log scale, the scatterplot shows an overall upward trend, but the relationship is not perfectly linear. The spread of the observations changes across the range of mass and radius, and several observations appear separated from the main pattern. These observations can be considered potential outliers and should be investigated rather than automatically removed.
```{r}
# Filter out missing values
pair2 <- exo_clean %>% 
  filter(!is.na(pl_bmasse) & !is.na(pl_rade))

# Calculate Pearson Correlation
cor_pair2 <- cor(pair2$pl_bmasse, pair2$pl_rade, method = "pearson")
cat("Correlation between Mass and Radius:", round(cor_pair2, 2), "\n")

# Simple Scatterplot
ggplot(pair2, aes(x = pl_bmasse, y = pl_rade)) +
  geom_point(alpha = 0.5, color = "darkgreen") +
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(method = "lm", color = "orange", se = FALSE) +
  labs(
    title = "Planet Mass vs. Planet Radius",
    x = "Planet Mass (Earth Mass, Log Scale)",
    y = "Planet Radius (Earth Radii, Log Scale)"
  ) +
  theme_minimal()
```


# 7. Distributional Analysis and QQ Plots

We study the equilibrium temperature of planets (`pl_eqt`, in Kelvin). It is a continuous numerical variable, so the Normal distribution is a natural distribution to compare it with. We are not saying that the variable is Normal, we only check if Normal is a reasonable description.

## 7.1 QQ Plots

```{r qq-overall}
eqt <- exo_clean$pl_eqt[!is.na(exo_clean$pl_eqt)]
length(eqt)

qqnorm(eqt,
       main = "Normal QQ Plot of Equilibrium Temperature",
       xlab = "Theoretical Normal Quantiles",
       ylab = "Observed Quantiles")
qqline(eqt)
```

What we see:

- The points follow the line only in the middle part.
- Away from the middle they leave the line and make a curve.
- The curve is smooth and not random, so the deviation is systematic (it is not just because of one or two points).
- The biggest deviation is in the upper tail: the real high temperatures are much higher than what a Normal distribution would give.
- The points at the top right (close to 4000 K) are far above the line, so they are potential unusual observations. We cannot say they are errors, because a QQ plot alone cannot tell us that.
- This curved shape is typical for a right-skewed distribution.

**Conclusion:** `pl_eqt` is not reasonably Normal.

## 7.2 Does the distribution change by discovery method?

We make one vector of temperatures for each method (without missing values) and check the sample sizes.

```{r groups}
transit <- exo_clean$pl_eqt[exo_clean$discoverymethod == "Transit" &
                              !is.na(exo_clean$pl_eqt)]

rv <- exo_clean$pl_eqt[exo_clean$discoverymethod == "Radial Velocity" &
                         !is.na(exo_clean$pl_eqt)]

imaging <- exo_clean$pl_eqt[exo_clean$discoverymethod == "Imaging" &
                              !is.na(exo_clean$pl_eqt)]

length(transit)
length(rv)
length(imaging)
```

Transit has over 17000 planets but Imaging has only 101, so we need to be more careful with the Imaging results.

```{r boxplot-groups}
boxplot(transit, rv, imaging,
        names = c("Transit", "Radial Velocity", "Imaging"),
        main = "Equilibrium Temperature by Discovery Method",
        xlab = "Discovery Method",
        ylab = "Equilibrium Temperature (K)")
```

The Imaging box is higher and much bigger than the other two. Transit has a lot of dots (potential outliers) on the top, Radial Velocity has only a few and Imaging has none.

Numbers for each group:

```{r summaries-groups}
# Transit
summary(transit)
sd(transit)
IQR(transit)
mean((transit - mean(transit))^3) / (mean((transit - mean(transit))^2))^1.5   # skewness

# Radial Velocity
summary(rv)
sd(rv)
IQR(rv)
mean((rv - mean(rv))^3) / (mean((rv - mean(rv))^2))^1.5   # skewness

# Imaging
summary(imaging)
sd(imaging)
IQR(imaging)
mean((imaging - mean(imaging))^3) / (mean((imaging - mean(imaging))^2))^1.5   # skewness
```


QQ plot for each group:

```{r qq-groups, fig.width=10, fig.height=4.5}
par(mfrow = c(1, 3))

qqnorm(transit, main = "Transit")
qqline(transit)

qqnorm(rv, main = "Radial Velocity")
qqline(rv)

qqnorm(imaging, main = "Imaging")
qqline(imaging)

par(mfrow = c(1, 1))
```

What we see:

- Transit and Radial Velocity curve upward away from the line, so they are right-skewed (not Normal).
- Imaging points are much closer to the line in the middle, so it is more symmetric. It has small deviations at the ends, but the sample is small (101), so we should not trust this too much.

### Answers to the questions of 7.2

**1. Can the distributions be assumed to be the same?**
No, they are clearly not the same.

**2. How do they differ?**

- **Location:** Imaging has the highest median (1500 K), then Transit (800 K), and Radial Velocity is the lowest (524 K). The means have the same order as the medians, so it is not because of a few extreme values.
- **Spread:** Imaging IQR is 1025 K, almost double of Transit (532 K) and Radial Velocity (517 K). Imaging also has the biggest standard deviation (683.9 K).
- **Skewness and shape:** Transit (1.23) and Radial Velocity (1.00) are right-skewed. Imaging (-0.02) is approximately symmetric.
- **Tails:** Transit has a very long upper tail with many potential outliers. Radial Velocity has few and Imaging has none.

**3. Are the differences big or small?**
Big. They differ in location, spread and shape at the same time.

**4. Is some category different from the others?**
Yes, Imaging: higher center, much bigger spread and a symmetric shape instead of right-skewed. Transit and Radial Velocity are more similar to each other.

This is only descriptive analysis. We see an association between discovery method and temperature, we do not say the method causes the temperature differences. Also Imaging has a small sample.

```