---
title: "CS 108 Statistics — Project 1"
date: "2026-09-27"
output:
  html_document:
    toc: true
    toc_depth: 2
  
---

```
{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 8, fig.height = 5)
```

# 2 Data Preparation
## 2.1 Data Import and Structure Examination
The dataset was imported into R from the NASA Exoplanet Archive Planetary Systems (PS) table. The dataset initially contains tens of thousands of rows and multiple columns capturing orbital, stellar, and physical planetary attributes. Inspecting the data structure using str() confirms that numerical variables (such as planet radius pl_rade, orbital period pl_orbper, mass pl_bmasse, and semi-major axis pl_orbsmax) are stored as double-precision floating-point numbers (num), while categorical identifiers (such as discovery method discoverymethod and stellar multiplicity sy_snum) were initially read as character or integer types.

```{r load-data, echo=TRUE}
library(readr)
library(ggplot2)
library(dplyr)

exo <- read_csv("ExoplanetsDataset.csv", comment = "#",  show_col_types = FALSE)

dim(exo)
names(exo)
str(exo)
```
## 2.2 Missing Values, Inconsistencies, and Unusual Observations
- **Missing Value Analysis**: The missing-value analysis showed that the completeness of variables differs substantially. Orbital period has fewer missing values than several physical properties such as planet mass and radius. Because missingness varies across variables, analyses involving these variables use only observations for which the required variables are available.
```{r}
key_vars <- c(
  "pl_rade",
  "pl_orbper",
  "pl_bmasse",
  "pl_orbsmax",
  "discoverymethod",
  "sy_snum"
)

missing_summary <- data.frame(
  Variable = key_vars,
  Missing = sapply(exo[key_vars], function(x) sum(is.na(x))),
  Missing_Percent = round(
    sapply(exo[key_vars], function(x) mean(is.na(x))) * 100, 2
  )
)

missing_summary
```
- **Data Integrity Checks**: Sanity checks confirmed that physical attributes contain no non-physical zero or negative values.
```{r}
sum(exo$pl_rade <= 0, na.rm = TRUE)
sum(exo$pl_orbper <= 0, na.rm = TRUE)
sum(exo$pl_bmasse <= 0, na.rm = TRUE)
sum(exo$pl_orbsmax <= 0, na.rm = TRUE)
```
- **Handling Duplicate Entries**: The primary Planetary Systems table contains multiple published parameter sets for single planets across different scientific papers. Filtering by default_flag == 1 isolates the single default, peer-reviewed parameter set for each distinct exoplanet, preventing duplicate weighting in subsequent EDA steps.
```{r}
sum(exo$default_flag == 1, na.rm = TRUE)
```


## 2.3 Data Type Conversions and Explicit Explanations
``` {r}

```


# 5. Numerical Variables by Categories

## 5.1 Planet Radius (pl_rade) by Discovery Method (discoverymethod)
 - **Differences in Location**: The median planet radius differs across the major discovery-method groups. The Transit group has a lower median radius than some of the other discovery methods. The median is useful for this comparison because planet radius has a skewed distribution and can contain extreme values.
 - **Differences in Variability and Shape**:The distributions also differ in their spread and shape. The Transit group has many observations concentrated at smaller planet radii, with a longer tail toward larger radii. The boxplot also shows differences in the IQR and the presence of potential outliers across discovery methods.
 To make the comparison readable, the four most frequent discovery methods were retained as separate categories and all remaining methods were combined into an "Other" category. This reduces the number of groups shown in the graph, but the "Other" category contains heterogeneous discovery methods and therefore should not be interpreted as a single detection technique.

```{r}
exo_clean <- exo %>% filter(default_flag == 1)

# ----------------------------------------------------------------------
#  Log Planet Radius (pl_rade) across Discovery Method (discoverymethod)
# ----------------------------------------------------------------------
# Categorical grouping: Top 4 most frequent discovery methods vs Other
top_methods <- exo_clean %>%
  count(discoverymethod) %>%
  top_n(4, n) %>%
  pull(discoverymethod)

exo_clean <- exo_clean %>%
  mutate(disc_method_cat = ifelse(discoverymethod %in% top_methods, discoverymethod, "Other"))

# Grouped Summary Table: Location and Spread
radius_by_method <- exo_clean %>%
  filter(!is.na(pl_rade)) %>%
  group_by(disc_method_cat) %>%
  summarise(
    Count = n(),
    Mean = mean(pl_rade, na.rm = TRUE),
    Median = median(pl_rade, na.rm = TRUE),
    SD = sd(pl_rade, na.rm = TRUE),
    IQR = IQR(pl_rade, na.rm = TRUE),
    Min = min(pl_rade, na.rm = TRUE),
    Max = max(pl_rade, na.rm = TRUE)
  )
radius_by_method


# Side-by-side Boxplot (Log10 scale due to skewness)
ggplot(exo_clean %>% filter(!is.na(pl_rade)), 
       aes(x = disc_method_cat, y = pl_rade, fill = disc_method_cat)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.3) +
  scale_y_log10() +
  labs(title = "Planet Radius Distribution by Discovery Method",
       x = "Discovery Method",
       y = "Planet Radius (Earth Radii, Log Scale)",
       fill = "Method") +
  theme_minimal() +
  theme(legend.position = "none")

Q1 <- quantile(exo_clean$pl_rade, 0.25, na.rm = TRUE)
Q3 <- quantile(exo_clean$pl_rade, 0.75, na.rm = TRUE)
IQR_value <- Q3 - Q1

outliers <- exo_clean %>%
  filter(
    pl_rade < Q1 - 1.5 * IQR_value |
    pl_rade > Q3 + 1.5 * IQR_value
  )

outliers %>%
  select(pl_name, pl_rade, discoverymethod)
```


## 5.2 Orbital Period (pl_orbper) across Stellar Multiplicity (sy_snum)
The distribution of orbital period was compared across planetary systems with different numbers of stars. The sy_snum variable was grouped into three categories: Single Star (sy_snum = 1), Binary Star System (sy_snum = 2), and Multi-Star (3+) (sy_snum ≥ 3). Only observations with available orbital-period values were included in this comparison.
### Differences in Location
The median orbital period is 10.6 days for Single Star systems, 12.8 days for Binary Star Systems, and 13.7 days for Multi-Star systems. Therefore, the typical orbital period is somewhat longer in systems with two or more stars.
The means are much larger than the medians in all three groups. For example, the mean for Binary Star Systems is 16,857 days, compared with a median of only 12.8 days. This large difference suggests that a relatively small number of planets with very long orbital periods strongly affect the mean. Therefore, the median is more useful than the mean for comparing the typical orbital period across these groups.
### Differences in Variability and Shape
There are substantial differences in variability between the groups. The IQR is 30.8 days for Single Star systems, 213 days for Binary Star Systems, and 252 days for Multi-Star systems. This shows that the middle 50% of orbital periods are much more spread out in the Binary and Multi-Star groups.
The standard deviations are also very large, especially for Binary Star Systems. The standard deviation is 147,179 days for Single Star systems, 278,352 days for Binary Star systems, and 12,515 days for Multi-Star systems. These large values indicate that extreme orbital-period observations have a strong effect on the standard deviation.

Overall, Most discovery-method groups have right-skewed distributions, as shown by the large differences between the means and medians. The boxplot and density plot on the logarithmic scale make differences in spread and the presence of extreme values easier to see. These results show an association between stellar multiplicity and the distribution of orbital periods, but they do not establish that the number of stars causes differences in orbital period.
```{r}
exo_clean <- exo_clean %>%
  mutate(snum_cat = factor(case_when(
    sy_snum == 1 ~ "Single Star",
    sy_snum == 2 ~ "Binary Star System",
    sy_snum >= 3 ~ "Multi-Star (3+)",
    TRUE ~ NA_character_
  )))

period_by_snum <- exo_clean %>%
  filter(!is.na(pl_orbper) & !is.na(snum_cat)) %>%
  group_by(snum_cat) %>%
  summarise(
    Count = n(),
    Mean = mean(pl_orbper, na.rm = TRUE),
    Median = median(pl_orbper, na.rm = TRUE),
    SD = sd(pl_orbper, na.rm = TRUE),
    IQR = IQR(pl_orbper, na.rm = TRUE)
  )
print(period_by_snum)

# Grouped Density Plot
ggplot(exo_clean %>% filter(!is.na(pl_orbper) & !is.na(snum_cat)), 
       aes(x = pl_orbper, fill = snum_cat)) +
  geom_density(alpha = 0.4) +
  scale_x_log10() +
  labs(title = "Density Distribution of Orbital Period by Stellar Multiplicity",
       x = "Orbital Period (Days, Log Scale)",
       y = "Density",
       fill = "Stellar System") +
  theme_minimal()

ggplot(
  exo_clean %>% filter(!is.na(pl_orbper), !is.na(snum_cat)),
  aes(x = snum_cat, y = pl_orbper)
) +
  geom_boxplot() +
  scale_y_log10() +
  labs(
    title = "Orbital Period by Stellar Multiplicity",
    x = "Stellar System",
    y = "Orbital Period (Days, Log Scale)"
  ) +
  theme_minimal()

```

# 6: Relationships Between Numerical Variables
## 6.1 Distance from Star (pl_orbsmax) vs. Orbital Period (pl_orbper)
- **Direction and Strength**: There is a strong positive relationship between a planet's semi-major axis and its orbital period. In general, planets with larger semi-major axes tend to have longer orbital periods. The Pearson correlation coefficient indicates a strong positive association between the two variables.
- **Form and Outliers**: On the logarithmic scales, the scatterplot shows a clear positive and approximately linear pattern.This indicates that the relationship is well described by a power-law-type pattern on the log-log scale. A small number of observations appear far from the main concentration of points, particularly at very large semi-major axes.

```{r}
# Semi-Major Axis vs. Orbital Period

# Filter out missing values, i.e complete-case analysis for this pair.
pair1 <- exo_clean %>% 
  filter(!is.na(pl_orbsmax) & !is.na(pl_orbper))

# Calculate Pearson Correlation
cor_pair1 <- cor(pair1$pl_orbsmax, pair1$pl_orbper, method = "pearson")
cat("Correlation between Distance and Period:", round(cor_pair1, 2), "\n")

# Simple Scatterplot
ggplot(pair1, aes(x = pl_orbsmax, y = pl_orbper)) +
  geom_point(alpha = 0.5, color = "steelblue") +
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  labs(
    title = "Orbital Period vs. Distance from Host Star",
    x = "Distance from Star (AU, Log Scale)",
    y = "Orbital Period (Days, Log Scale)"
  ) +
  theme_minimal()
```

## 6.2 Planet Mass (pl_bmasse) vs. Planet Radius (pl_rade)
- **Direction and Strength**: There is a positive relationship between planet mass and planet radius. In general, planets with larger masses tend to have larger radii, although the points show considerable variation around the overall trend.
- **Form and Outliers**: On the log-log scale, the scatterplot shows an overall upward trend, but the relationship is not perfectly linear. The spread of the observations changes across the range of mass and radius, and several observations appear separated from the main pattern. These observations can be considered potential outliers and should be investigated rather than automatically removed.
```{r}
# Filter out missing values
pair2 <- exo_clean %>% 
  filter(!is.na(pl_bmasse) & !is.na(pl_rade))

# Calculate Pearson Correlation
cor_pair2 <- cor(pair2$pl_bmasse, pair2$pl_rade, method = "pearson")
cat("Correlation between Mass and Radius:", round(cor_pair2, 2), "\n")

# Simple Scatterplot
ggplot(pair2, aes(x = pl_bmasse, y = pl_rade)) +
  geom_point(alpha = 0.5, color = "darkgreen") +
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(method = "lm", color = "orange", se = FALSE) +
  labs(
    title = "Planet Mass vs. Planet Radius",
    x = "Planet Mass (Earth Mass, Log Scale)",
    y = "Planet Radius (Earth Radii, Log Scale)"
  ) +
  theme_minimal()
```





# 7. Distributional Analysis and QQ Plots

We study **planet equilibrium temperature (`pl_eqt`)**, measured in Kelvin.

Our questions are:

1. Can equilibrium temperature be reasonably described by a Normal distribution?
2. Does its distribution change by discovery method?

## 7.1 Overall Distribution

First, we remove missing values.

```{r point7-data}
eqt <- exo_clean$pl_eqt[!is.na(exo_clean$pl_eqt)]

length(eqt)
```

We calculate the main descriptive statistics.

```{r point7-statistics}
eqt_mean <- mean(eqt)
eqt_median <- median(eqt)
eqt_min <- min(eqt)
eqt_max <- max(eqt)
eqt_range <- eqt_max - eqt_min
eqt_q1 <- quantile(eqt, 0.25)
eqt_q3 <- quantile(eqt, 0.75)
eqt_iqr <- IQR(eqt)
eqt_sd <- sd(eqt)

eqt_skewness <- mean((eqt - eqt_mean)^3) /
  (mean((eqt - eqt_mean)^2))^1.5

cat("Number of observations:", length(eqt), "\n")
cat("Mean:", round(eqt_mean, 2), "K\n")
cat("Median:", round(eqt_median, 2), "K\n")
cat("Minimum:", eqt_min, "K\n")
cat("Maximum:", eqt_max, "K\n")
cat("Range:", round(eqt_range, 2), "K\n")
cat("Q1:", round(eqt_q1, 2), "K\n")
cat("Q3:", round(eqt_q3, 2), "K\n")
cat("IQR:", round(eqt_iqr, 2), "K\n")
cat("Standard deviation:", round(eqt_sd, 2), "K\n")
cat("Skewness:", round(eqt_skewness, 2), "\n")
```

The mean is higher than the median and the skewness is positive, suggesting a right-skewed distribution.

### Histogram

```{r point7-histogram, fig.width=9, fig.height=5}
hist(
  eqt,
  breaks = 40,
  col = "#5B8FF9",
  border = "white",
  main = "Distribution of Planet Equilibrium Temperature",
  xlab = "Equilibrium Temperature (K)",
  ylab = "Number of Planets"
)

abline(
  v = eqt_mean,
  col = "#E74C3C",
  lwd = 3,
  lty = 2
)

abline(
  v = eqt_median,
  col = "#2ECC71",
  lwd = 3,
  lty = 2
)

legend(
  "topright",
  legend = c(
    paste("Mean =", round(eqt_mean, 1), "K"),
    paste("Median =", round(eqt_median, 1), "K")
  ),
  col = c("#E74C3C", "#2ECC71"),
  lty = 2,
  lwd = 3,
  bty = "n"
)
```

The histogram has a long upper tail, so the distribution is strongly right-skewed.

## 7.2 Normal Distribution and QQ Plot

We use the Normal distribution as a theoretical reference to see whether it is a reasonable description of the data.

```{r point7-overall-qq, fig.width=8, fig.height=6}
qqnorm(
  eqt,
  main = "Normal QQ Plot of Planet Equilibrium Temperature",
  xlab = "Theoretical Normal Quantiles",
  ylab = "Observed Quantiles",
  pch = 16,
  col = "#5B8FF9"
)

qqline(
  eqt,
  col = "#E74C3C",
  lwd = 3
)
```

The points are closer to the line in the middle but deviate systematically in the tails, especially the upper tail.

Therefore, equilibrium temperature is **not reasonably described by a Normal distribution**.

The extreme upper observations are potentially unusual, but we do not automatically consider them errors.

## 7.3 Distribution by Discovery Method

We compare:

- Transit
- Radial Velocity
- Imaging

```{r point7-groups}
transit <- exo_clean$pl_eqt[
  exo_clean$discoverymethod == "Transit" &
    !is.na(exo_clean$pl_eqt)
]

rv <- exo_clean$pl_eqt[
  exo_clean$discoverymethod == "Radial Velocity" &
    !is.na(exo_clean$pl_eqt)
]

imaging <- exo_clean$pl_eqt[
  exo_clean$discoverymethod == "Imaging" &
    !is.na(exo_clean$pl_eqt)
]

cat("Transit:", length(transit), "\n")
cat("Radial Velocity:", length(rv), "\n")
cat("Imaging:", length(imaging), "\n")
```

Transit has more than 17,000 observations, while Imaging has about 101, so the Imaging results should be interpreted more carefully.

### Boxplot

```{r point7-boxplot, fig.width=9, fig.height=6}
boxplot(
  transit,
  rv,
  imaging,
  names = c("Transit", "Radial Velocity", "Imaging"),
  col = c("#5B8FF9", "#9B59B6", "#2ECC71"),
  border = "black",
  main = "Equilibrium Temperature by Discovery Method",
  xlab = "Discovery Method",
  ylab = "Equilibrium Temperature (K)",
  las = 1
)
```

Imaging has the highest center and the largest spread.

Transit and Radial Velocity have lower centers and more right-skewed distributions.

### Numerical Summary

```{r point7-group-statistics}
transit_mean <- mean(transit)
transit_median <- median(transit)
transit_sd <- sd(transit)
transit_iqr <- IQR(transit)

transit_skew <- mean((transit - transit_mean)^3) /
  (mean((transit - transit_mean)^2))^1.5

rv_mean <- mean(rv)
rv_median <- median(rv)
rv_sd <- sd(rv)
rv_iqr <- IQR(rv)

rv_skew <- mean((rv - rv_mean)^3) /
  (mean((rv - rv_mean)^2))^1.5

imaging_mean <- mean(imaging)
imaging_median <- median(imaging)
imaging_sd <- sd(imaging)
imaging_iqr <- IQR(imaging)

imaging_skew <- mean((imaging - imaging_mean)^3) /
  (mean((imaging - imaging_mean)^2))^1.5

group_summary <- data.frame(
  Method = c(
    "Transit",
    "Radial Velocity",
    "Imaging"
  ),
  Count = c(
    length(transit),
    length(rv),
    length(imaging)
  ),
  Mean = c(
    transit_mean,
    rv_mean,
    imaging_mean
  ),
  Median = c(
    transit_median,
    rv_median,
    imaging_median
  ),
  SD = c(
    transit_sd,
    rv_sd,
    imaging_sd
  ),
  IQR = c(
    transit_iqr,
    rv_iqr,
    imaging_iqr
  ),
  Skewness = c(
    transit_skew,
    rv_skew,
    imaging_skew
  )
)

knitr::kable(
  group_summary,
  digits = 2,
  caption = "Equilibrium Temperature by Discovery Method"
)
```

Imaging has the highest median and largest IQR.

Transit and Radial Velocity have positive skewness, while Imaging has skewness close to zero.

## 7.4 QQ Plots by Discovery Method

```{r point7-group-qq, fig.width=12, fig.height=5}
par(mfrow = c(1, 3))

qqnorm(
  transit,
  main = "Transit",
  xlab = "Theoretical Normal Quantiles",
  ylab = "Observed Quantiles",
  pch = 16,
  col = "#5B8FF9"
)

qqline(
  transit,
  col = "#E74C3C",
  lwd = 2
)

qqnorm(
  rv,
  main = "Radial Velocity",
  xlab = "Theoretical Normal Quantiles",
  ylab = "Observed Quantiles",
  pch = 16,
  col = "#9B59B6"
)

qqline(
  rv,
  col = "#E74C3C",
  lwd = 2
)

qqnorm(
  imaging,
  main = "Imaging",
  xlab = "Theoretical Normal Quantiles",
  ylab = "Observed Quantiles",
  pch = 16,
  col = "#2ECC71"
)

qqline(
  imaging,
  col = "#E74C3C",
  lwd = 2
)

par(mfrow = c(1, 1))
```

Transit and Radial Velocity show stronger curvature and right-skewness.

Imaging is more symmetric and closer to a linear pattern in the middle, but its sample size is much smaller.

## 7.5 Final Conclusion

The three discovery methods have clearly different equilibrium-temperature distributions.

- **Location:** Imaging has the highest median.
- **Spread:** Imaging has the largest IQR.
- **Skewness:** Transit and Radial Velocity are right-skewed.
- **Tails:** Transit has a particularly long upper tail.
- **Shape:** Imaging is more symmetric.

Overall, the differences are substantial rather than minor.

The results show an association between discovery method and equilibrium temperature, but they do not show that discovery method causes the differences.

```
