---
title: "CS 108 Statistics — Project 1"
date: "2026-09-27"
output:
  html_document:
    toc: true
    toc_depth: 2
  
---

```
{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 8, fig.height = 5)
```

# 2 Data Preparation
## 2.1 Data Import and Structure Examination
The dataset was imported into R from the NASA Exoplanet Archive Planetary Systems (PS) table. The dataset initially contains tens of thousands of rows and multiple columns capturing orbital, stellar, and physical planetary attributes. Inspecting the data structure using str() confirms that numerical variables (such as planet radius pl_rade, orbital period pl_orbper, mass pl_bmasse, and semi-major axis pl_orbsmax) are stored as double-precision floating-point numbers (num), while categorical identifiers (such as discovery method discoverymethod and stellar multiplicity sy_snum) were initially read as character or integer types.

```{r load-data, echo=TRUE}
library(readr)
library(ggplot2)
library(dplyr)

exo <- read_csv("ExoplanetsDataset.csv", comment = "#",  show_col_types = FALSE)

dim(exo)
names(exo)
str(exo)
```
## 2.2 Missing Values, Inconsistencies, and Unusual Observations
- **Missing Value Analysis**: The missing-value analysis showed that the completeness of variables differs substantially. Orbital period has fewer missing values than several physical properties such as planet mass and radius. Because missingness varies across variables, analyses involving these variables use only observations for which the required variables are available.
```{r}
key_vars <- c(
  "pl_rade",
  "pl_orbper",
  "pl_bmasse",
  "pl_orbsmax",
  "discoverymethod",
  "sy_snum"
)

missing_summary <- data.frame(
  Variable = key_vars,
  Missing = sapply(exo[key_vars], function(x) sum(is.na(x))),
  Missing_Percent = round(
    sapply(exo[key_vars], function(x) mean(is.na(x))) * 100, 2
  )
)

missing_summary
```
- **Data Integrity Checks**: Sanity checks confirmed that physical attributes contain no non-physical zero or negative values.
```{r}
sum(exo$pl_rade <= 0, na.rm = TRUE)
sum(exo$pl_orbper <= 0, na.rm = TRUE)
sum(exo$pl_bmasse <= 0, na.rm = TRUE)
sum(exo$pl_orbsmax <= 0, na.rm = TRUE)
```
- **Handling Duplicate Entries**: The primary Planetary Systems table contains multiple published parameter sets for single planets across different scientific papers. Filtering by default_flag == 1 isolates the single default, peer-reviewed parameter set for each distinct exoplanet, preventing duplicate weighting in subsequent EDA steps.
```{r}
sum(exo$default_flag == 1, na.rm = TRUE)
```


## 2.3 Data Type Conversions and Explicit Explanations
``` {r}

```


# 5. Numerical Variables by Categories

## 5.1 Planet Radius (pl_rade) by Discovery Method (discoverymethod)
 - **Differences in Location**: **Transit** planets have a lower median radius than the other major discovery-method groups. Transit surveys excel at finding Earth-to-Neptune sized candidates blocking faint stellar light, whereas Direct Imaging detects young, self-luminous giant planets.
 - **Differences in Variability and Shape**: The Transit radius distribution is positively skewed with a prominent concentration around Super-Earths/Sub-Neptunes ($1.5 - 3.5\ R_\oplus$) and a thin right-side tail. Radial Velocity exhibits broader spread (IQR) across planet radii due to mass-based detection filters.
 To make the comparison readable, the four most frequent discovery methods were retained as separate categories and all remaining methods were combined into an "Other" category. This reduces the number of groups shown in the graph, but the "Other" category contains heterogeneous discovery methods and therefore should not be interpreted as a single detection technique.

```{r}
exo_clean <- exo %>% filter(default_flag == 1)

# ----------------------------------------------------------------------
#  Log Planet Radius (pl_rade) across Discovery Method (discoverymethod)
# ----------------------------------------------------------------------
# Categorical grouping: Top 4 most frequent discovery methods vs Other
top_methods <- exo_clean %>%
  count(discoverymethod) %>%
  top_n(4, n) %>%
  pull(discoverymethod)

exo_clean <- exo_clean %>%
  mutate(disc_method_cat = ifelse(discoverymethod %in% top_methods, discoverymethod, "Other"))

# Grouped Summary Table: Location and Spread
radius_by_method <- exo_clean %>%
  filter(!is.na(pl_rade)) %>%
  group_by(disc_method_cat) %>%
  summarise(
    Count = n(),
    Mean = mean(pl_rade, na.rm = TRUE),
    Median = median(pl_rade, na.rm = TRUE),
    SD = sd(pl_rade, na.rm = TRUE),
    IQR = IQR(pl_rade, na.rm = TRUE),
    Min = min(pl_rade, na.rm = TRUE),
    Max = max(pl_rade, na.rm = TRUE)
  )
radius_by_method


# Side-by-side Boxplot (Log10 scale due to skewness)
ggplot(exo_clean %>% filter(!is.na(pl_rade)), 
       aes(x = disc_method_cat, y = pl_rade, fill = disc_method_cat)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.3) +
  scale_y_log10() +
  labs(title = "Planet Radius Distribution by Discovery Method",
       x = "Discovery Method",
       y = "Planet Radius (Earth Radii, Log Scale)",
       fill = "Method") +
  theme_minimal() +
  theme(legend.position = "none")

Q1 <- quantile(exo_clean$pl_rade, 0.25, na.rm = TRUE)
Q3 <- quantile(exo_clean$pl_rade, 0.75, na.rm = TRUE)
IQR_value <- Q3 - Q1

outliers <- exo_clean %>%
  filter(
    pl_rade < Q1 - 1.5 * IQR_value |
    pl_rade > Q3 + 1.5 * IQR_value
  )

outliers %>%
  select(pl_name, pl_rade, discoverymethod)
```


## 5.2 Orbital Period (pl_orbper) across Stellar Multiplicity (sy_snum)
The distribution of orbital period was compared across planetary systems with different numbers of stars. The sy_snum variable was grouped into three categories: Single Star (sy_snum = 1), Binary Star System (sy_snum = 2), and Multi-Star (3+) (sy_snum ≥ 3). Only observations with available orbital-period values were included in this comparison.

### Differences in Location

The median orbital period is useful for comparing the typical planet in each group because orbital periods have a strongly right-skewed distribution. The group medians show whether planets in single-star, binary-star, and multi-star systems tend to have different typical orbital periods. The means are also reported, but they can be influenced by planets with very long orbital periods. Therefore, the median provides a more representative measure of the center of the distributions.

### Differences in Variability and Shape

The groups also differ in their variability and distributional shape. The standard deviation and IQR describe the spread of orbital periods within each stellar-system category. The density plot, shown on a logarithmic x-axis, makes it easier to compare orbital periods across their wide range of values. The distributions are right-skewed, with most planets having relatively shorter orbital periods and a smaller number having much longer periods. The boxplots also help identify potential outliers with unusually short or long orbital periods.

Overall, the comparison shows that stellar multiplicity is associated with differences in the distribution of orbital periods. However, these results describe patterns and associations in the dataset and do not establish that the number of stars causes planets to have particular orbital periods.

```{r}
exo_clean <- exo_clean %>%
  mutate(snum_cat = factor(case_when(
    sy_snum == 1 ~ "Single Star",
    sy_snum == 2 ~ "Binary Star System",
    sy_snum >= 3 ~ "Multi-Star (3+)",
    TRUE ~ NA_character_
  )))

period_by_snum <- exo_clean %>%
  filter(!is.na(pl_orbper) & !is.na(snum_cat)) %>%
  group_by(snum_cat) %>%
  summarise(
    Count = n(),
    Mean = mean(pl_orbper, na.rm = TRUE),
    Median = median(pl_orbper, na.rm = TRUE),
    SD = sd(pl_orbper, na.rm = TRUE),
    IQR = IQR(pl_orbper, na.rm = TRUE)
  )
print(period_by_snum)

# Grouped Density Plot
ggplot(exo_clean %>% filter(!is.na(pl_orbper) & !is.na(snum_cat)), 
       aes(x = pl_orbper, fill = snum_cat)) +
  geom_density(alpha = 0.4) +
  scale_x_log10() +
  labs(title = "Density Distribution of Orbital Period by Stellar Multiplicity",
       x = "Orbital Period (Days, Log Scale)",
       y = "Density",
       fill = "Stellar System") +
  theme_minimal()

ggplot(
  exo_clean %>% filter(!is.na(pl_orbper), !is.na(snum_cat)),
  aes(x = snum_cat, y = pl_orbper)
) +
  geom_boxplot() +
  scale_y_log10() +
  labs(
    title = "Orbital Period by Stellar Multiplicity",
    x = "Stellar System",
    y = "Orbital Period (Days, Log Scale)"
  ) +
  theme_minimal()

```

# 6: Relationships Between Numerical Variables
## 6.1 Distance from Star (pl_orbsmax) vs. Orbital Period (pl_orbper)
- **Direction and Strength**: There is a strong, positive relationship between a planet's distance from its star and its orbital period. As distance increases, orbital period increases. The calculated Pearson correlation coefficient is positive and indicates a strong association once plotted on a logarithmic scale.
- **Form and Outliers**: The scatter plot shows a clear straight line when using logarithmic scales for both axes. This shows that the two variables follow a steady proportional pattern (Kepler's Third Law: planets further from their star take longer to complete an orbit). Outliers are minimal, located mostly at extreme distances over $100\text{ AU}$.

```{r}
# Semi-Major Axis vs. Orbital Period

# Filter out missing values, i.e complete-case analysis for this pair.
pair1 <- exo_clean %>% 
  filter(!is.na(pl_orbsmax) & !is.na(pl_orbper))

# Calculate Pearson Correlation
cor_pair1 <- cor(pair1$pl_orbsmax, pair1$pl_orbper, method = "pearson")
cat("Correlation between Distance and Period:", round(cor_pair1, 2), "\n")

# Simple Scatterplot
ggplot(pair1, aes(x = pl_orbsmax, y = pl_orbper)) +
  geom_point(alpha = 0.5, color = "steelblue") +
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  labs(
    title = "Orbital Period vs. Distance from Host Star",
    x = "Distance from Star (AU, Log Scale)",
    y = "Orbital Period (Days, Log Scale)"
  ) +
  theme_minimal()
```

## 6.2 Planet Mass (pl_bmasse) vs. Planet Radius (pl_rade)
- **Direction and Strength**: There is a positive relationship between planet mass and planet radius. There is a positive association between planet mass and radius: planets with larger masses tend, on average, to have larger radii.
- **Form and Outliers**: On log-log scales, the scatterplot exhibits an upward trend for smaller terrestrial planets and super-Earths. However, for very high-mass gas giants, the curve flattens out—meaning heavy gas planets get denser rather than larger because gravity compresses the gas. Outliers primarily occur among extremely large, low-mass "puffy" gas planets.
```{r}
# Filter out missing values
pair2 <- exo_clean %>% 
  filter(!is.na(pl_bmasse) & !is.na(pl_rade))

# Calculate Pearson Correlation
cor_pair2 <- cor(pair2$pl_bmasse, pair2$pl_rade, method = "pearson")
cat("Correlation between Mass and Radius:", round(cor_pair2, 2), "\n")

# Simple Scatterplot
ggplot(pair2, aes(x = pl_bmasse, y = pl_rade)) +
  geom_point(alpha = 0.5, color = "darkgreen") +
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(method = "lm", color = "orange", se = FALSE) +
  labs(
    title = "Planet Mass vs. Planet Radius",
    x = "Planet Mass (Earth Mass, Log Scale)",
    y = "Planet Radius (Earth Radii, Log Scale)"
  ) +
  theme_minimal()
```