---
title: "CS 108 Statistics — Project 1"
date: "2026-09-27"
output:
  html_document:
    toc: true
    toc_depth: 2
  
---

```
{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 8, fig.height = 5)
```

# 2 Data Preparation
## 2.1 Data Import and Structure Examination
The dataset was imported into R from the NASA Exoplanet Archive Planetary Systems (PS) table. The dataset initially contains tens of thousands of rows and multiple columns capturing orbital, stellar, and physical planetary attributes. Inspecting the data structure using str() confirms that numerical variables (such as planet radius pl_rade, orbital period pl_orbper, mass pl_bmasse, and semi-major axis pl_orbsmax) are stored as double-precision floating-point numbers (num), while categorical identifiers (such as discovery method discoverymethod and stellar multiplicity sy_snum) were initially read as character or integer types.

```{r load-data, echo=TRUE}
library(readr)
library(ggplot2)
library(dplyr)
library(tidyr)


exo <- read_csv("ExoplanetsDataset.csv", comment = "#",  show_col_types = FALSE)

dim(exo)
names(exo)
str(exo)
```
## 2.2 Missing Values, Inconsistencies, and Unusual Observations
- **Missing Value Analysis**: The missing-value analysis showed that the completeness of variables differs substantially. Orbital period has fewer missing values than several physical properties such as planet mass and radius. Because missingness varies across variables, analyses involving these variables use only observations for which the required variables are available.
```{r}
key_vars <- c(
  "pl_rade",
  "pl_orbper",
  "pl_bmasse",
  "pl_orbsmax",
  "discoverymethod",
  "sy_snum"
)

missing_summary <- data.frame(
  Variable = key_vars,
  Missing = sapply(exo[key_vars], function(x) sum(is.na(x))),
  Missing_Percent = round(
    sapply(exo[key_vars], function(x) mean(is.na(x))) * 100, 2
  )
)

missing_summary
```
- **Data Integrity Checks**: Sanity checks confirmed that physical attributes contain no non-physical zero or negative values.
```{r}
sum(exo$pl_rade <= 0, na.rm = TRUE)
sum(exo$pl_orbper <= 0, na.rm = TRUE)
sum(exo$pl_bmasse <= 0, na.rm = TRUE)
sum(exo$pl_orbsmax <= 0, na.rm = TRUE)
```
- **Handling Duplicate Entries**: The primary Planetary Systems table contains multiple published parameter sets for single planets across different scientific papers. Filtering by default_flag == 1 isolates the single default, peer-reviewed parameter set for each distinct exoplanet, preventing duplicate weighting in subsequent EDA steps.
```{r}
sum(exo$default_flag == 1, na.rm = TRUE)
```


## 2.3 Data Type Conversions and Explicit Explanations

Categorical variables were stored as character or integer types, so they were converted to factors. Two derived categorical variables were also created. No observations were removed in this step.

- `discoverymethod` -> factor.
- `disc_method_cat`: the four most frequent discovery methods are kept and all others are merged into "Other". This makes plots readable, but "Other" mixes several detection techniques and must not be read as one method.
- `sy_snum` -> `snum_cat` with levels Single Star (1), Binary Star System (2), Multi-Star (3+). The 3+ levels are merged because they are rare.
- `disc_year` -> `disc_era` (four time periods), so the discovery year can be treated as a categorical variable.
- `ttv_flag`, `pl_controv_flag` (0/1) -> factors with labels "No"/"Yes".

```{r conversions}
exo_clean <- exo %>% filter(default_flag == 1)   # one parameter set per planet

top_methods <- exo_clean %>%
  count(discoverymethod) %>%
  slice_max(n, n = 4, with_ties = FALSE) %>%
  pull(discoverymethod)

exo_clean <- exo_clean %>%
  mutate(
    discoverymethod = factor(discoverymethod),
    disc_method_cat = factor(ifelse(discoverymethod %in% top_methods,
                                    as.character(discoverymethod), "Other")),
    snum_cat = factor(case_when(
      sy_snum == 1 ~ "Single Star",
      sy_snum == 2 ~ "Binary Star System",
      sy_snum >= 3 ~ "Multi-Star (3+)"),
      levels = c("Single Star", "Binary Star System", "Multi-Star (3+)")),
    disc_era = cut(disc_year,
                   breaks = c(-Inf, 2009, 2014, 2019, Inf),
                   labels = c("Before 2010", "2010-2014", "2015-2019", "2020 onward")),
    ttv_flag = factor(ttv_flag, levels = c(0, 1), labels = c("No", "Yes")),
    pl_controv_flag = factor(pl_controv_flag, levels = c(0, 1), labels = c("No", "Yes"))
  )

nrow(exo_clean)   # number of distinct planets analysed
```

# 3. Analysis of Categorical Variables

The important categorical variables are the discovery method, the number of stars in the system, the discovery period, and the transit-timing-variation (TTV) flag. Each unit of observation is one planet (default parameter set).

## 3.1 Discovery Method (`discoverymethod`)

```{r cat-method-table}
tab_method <- table(exo_clean$discoverymethod)
method_df <- data.frame(
  Method = names(tab_method),
  Frequency = as.integer(tab_method),
  Relative_Freq = round(as.numeric(prop.table(tab_method)), 4),
  Percent = round(100 * as.numeric(prop.table(tab_method)), 2)
) %>% arrange(desc(Frequency))

knitr::kable(method_df, caption = "Frequency and relative frequency of discovery methods")
```

```{r cat-method-plot}
ggplot(method_df, aes(x = reorder(Method, Frequency), y = Frequency)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = paste0(Percent, "%")), hjust = -0.1, size = 3) +
  coord_flip() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(title = "Number of Planets by Discovery Method",
       x = "Discovery Method", y = "Number of Planets") +
  theme_minimal()
```

**Interpretation.** The distribution is far from balanced. The Transit method is by far the most common and accounts for 73.96% of all planets, followed by Radial Velocity 18.85%. Several methods (for example Microlensing, Imaging, Transit Timing Variations, Eclipse Timing Variations, Orbital Brightness Modulation, Astrometry, Pulsar Timing, Pulsation Timing Variations, Disk Kinematics) are rare, each representing only a few planets or a small fraction of the data. Such sparse categories are potentially problematic: summaries based on a handful of planets are unstable, and tables or tests involving them are unreliable. For this reason, the four most frequent methods are kept as separate categories and all others are combined into "Other" in later sections.

```{r cat-method-cat}
tab_cat <- table(exo_clean$disc_method_cat)
knitr::kable(
  data.frame(Category = names(tab_cat),
             Frequency = as.integer(tab_cat),
             Percent = round(100 * as.numeric(prop.table(tab_cat)), 2)),
  caption = "Discovery method after grouping (top 4 + Other)")
```

## 3.2 Number of Stars in the System (`snum_cat`)

```{r cat-snum}
tab_snum <- table(exo_clean$snum_cat, useNA = "ifany")
snum_df <- data.frame(
  Category = names(tab_snum),
  Frequency = as.integer(tab_snum),
  Percent = round(100 * as.numeric(prop.table(tab_snum)), 2)
)
knitr::kable(snum_df, caption = "Stellar multiplicity of the host systems")

ggplot(snum_df, aes(x = Category, y = Frequency, fill = Category)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = paste0(Percent, "%")), vjust = -0.3, size = 3.5) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  labs(title = "Planets by Stellar Multiplicity of the Host System",
       x = "Stellar System", y = "Number of Planets") +
  theme_minimal()
```

**Interpretation.** Most planets in the archive orbit single stars, 90.94%, while binary systems make up 7.89% and systems with three or more stars only 1.18%. The Multi-Star category is rare, so any comparison involving it is less reliable. Part of this imbalance may reflect observational selection: surveys often favour single stars because companion stars complicate detection.

## 3.3 Discovery Period (`disc_era`)

```{r cat-era}
tab_era <- table(exo_clean$disc_era, useNA = "ifany")
era_df <- data.frame(
  Period = names(tab_era),
  Frequency = as.integer(tab_era),
  Percent = round(100 * as.numeric(prop.table(tab_era)), 2)
)
knitr::kable(era_df, caption = "Planets by discovery period")

ggplot(era_df, aes(x = factor(Period, levels = levels(exo_clean$disc_era)), y = Frequency)) +
  geom_col(fill = "darkorange") +
  geom_text(aes(label = paste0(Percent, "%")), vjust = -0.3, size = 3.5) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  labs(title = "Number of Planets by Discovery Period",
       x = "Discovery Period", y = "Number of Planets") +
  theme_minimal()
```

**Interpretation.** The number of discoveries increased sharply over time: 93.64% of planets were found after 2010, compared with only 6.36% before 2010. The distribution is therefore strongly unbalanced toward recent years, which reflects the growth of large survey missions.

## 3.4 Transit Timing Variations (`ttv_flag`)

```{r cat-ttv}
tab_ttv <- table(exo_clean$ttv_flag, useNA = "ifany")
knitr::kable(
  data.frame(TTV = names(tab_ttv),
             Frequency = as.integer(tab_ttv),
             Percent = round(100 * as.numeric(prop.table(tab_ttv)), 2)),
  caption = "Planets with detected transit timing variations")
```

**Interpretation.** Only 7.76% planets show TTVs, so this is a heavily imbalanced binary variable. It is mainly useful as a supplementary variable.

## 3.5 Interaction Between Categorical Variables

### Pair 1: Discovery Method and Stellar Multiplicity

*Question: does the distribution of stellar multiplicity differ between discovery methods?*

```{r pair-method-snum}
ct1 <- table(exo_clean$disc_method_cat, exo_clean$snum_cat)
addmargins(ct1)

# Row proportions: distribution of stellar multiplicity within each method
round(100 * prop.table(ct1, margin = 1), 1)
```

```{r pair-method-snum-plot, fig.width=9, fig.height=5}
df1 <- exo_clean %>% filter(!is.na(snum_cat))

ggplot(df1, aes(x = disc_method_cat, fill = snum_cat)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Stellar Multiplicity within Each Discovery Method",
       x = "Discovery Method", y = "Proportion of Planets", fill = "Stellar System") +
  theme_minimal()

ggplot(df1, aes(x = disc_method_cat, fill = snum_cat)) +
  geom_bar(position = "dodge") +
  labs(title = "Counts by Discovery Method and Stellar Multiplicity",
       x = "Discovery Method", y = "Number of Planets", fill = "Stellar System") +
  theme_minimal()
```

**Interpretation.** In every method, single-star hosts form the largest group, but the proportions differ. For example, transit method has approximately 6% planets in binary systems, compared with approximately 18%% for Imaging. This indicates an association between the discovery method and the stellar multiplicity of the host. The share of Multi-Star systems is small everywhere. The counts plot should be used together with the proportion plot, since the stacked proportion bars hide the large differences in group sizes. This is an association only; it likely reflects how surveys select their target stars, not an effect of one variable on the other.

### Pair 2: Discovery Method and Discovery Period

*Question: has the mix of discovery methods changed over time?*

```{r pair-method-era}
ct2 <- table(exo_clean$disc_era, exo_clean$disc_method_cat)
addmargins(ct2)

# Row proportions: method mix within each period
round(100 * prop.table(ct2, margin = 1), 1)
```

```{r pair-method-era-plot, fig.width=9, fig.height=5}
df2 <- exo_clean %>% filter(!is.na(disc_era))

ggplot(df2, aes(x = disc_era, fill = disc_method_cat)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Discovery Method Mix by Discovery Period",
       x = "Discovery Period", y = "Proportion of Planets", fill = "Method") +
  theme_minimal()

ggplot(df2, aes(x = disc_era, fill = disc_method_cat)) +
  geom_bar(position = "dodge") +
  labs(title = "Number of Planets by Period and Discovery Method",
       x = "Discovery Period", y = "Number of Planets", fill = "Method") +
  theme_minimal()
```

**Interpretation.** The method mix changes clearly over time. In the earliest period, Radical Velocity dominates, whereas in later periods Transit becomes the dominant method. The distribution of one variable therefore depends strongly on the category of the other, so the two variables are clearly associated. A plausible explanation is the launch of dedicated transit surveys, but the data alone only show the association, not the reason.

# 4. Analysis of Numerical Variables

The key numerical variables are planet radius (`pl_rade`, Earth radii), planet mass (`pl_bmasse`, Earth masses), orbital period (`pl_orbper`, days), semi-major axis (`pl_orbsmax`, AU), and equilibrium temperature (`pl_eqt`, K). Missing values are excluded separately for each variable (available-case analysis), so the sample size differs between variables.

## 4.1 Descriptive Statistics

```{r num-summary}
num_vars <- c(
  pl_rade    = "Planet Radius (Earth radii)",
  pl_bmasse  = "Planet Mass (Earth masses)",
  pl_orbper  = "Orbital Period (days)",
  pl_orbsmax = "Semi-Major Axis (AU)",
  pl_eqt     = "Equilibrium Temperature (K)"
)

skew <- function(x) mean((x - mean(x))^3) / (mean((x - mean(x))^2))^1.5

num_summary <- function(x) {
  x <- x[!is.na(x)]
  q <- quantile(x, c(0.25, 0.75))
  fence_lo <- q[1] - 1.5 * IQR(x)
  fence_hi <- q[2] + 1.5 * IQR(x)
  data.frame(
    n = length(x), Mean = mean(x), Median = median(x),
    Min = min(x), Max = max(x), Range = max(x) - min(x),
    Q1 = q[[1]], Q3 = q[[2]], IQR = IQR(x), SD = sd(x),
    Outliers = sum(x < fence_lo | x > fence_hi),
    Outliers_pct = 100 * mean(x < fence_lo | x > fence_hi)
  )
}

summary_table <- do.call(rbind, lapply(names(num_vars), function(v) {
  cbind(Variable = num_vars[[v]], num_summary(exo_clean[[v]]))
}))

knitr::kable(summary_table, digits = 2, row.names = FALSE,
             caption = "Descriptive statistics of the key numerical variables")
```

*Note: the coefficient of variation (CV = SD / mean) is meaningful here because all variables are strictly positive ratio-scale quantities. "Outliers" is the number of observations outside the boxplot fences Q1 - 1.5 IQR and Q3 + 1.5 IQR.*

## 4.2 Distributions

For each variable, the histogram on the original scale, the histogram on the log10 scale, and a boxplot on the log10 scale are shown. The log scale is used because the variables span several orders of magnitude.

```{r num-plots, fig.width=12, fig.height=4}
for (v in names(num_vars)) {
  x <- exo_clean[[v]]
  x <- x[!is.na(x)]
  par(mfrow = c(1, 3))
  hist(x, breaks = 40, col = "#5B8FF9", border = "white",
       main = paste("Histogram:", num_vars[[v]]), xlab = num_vars[[v]])
  abline(v = c(mean(x), median(x)), col = c("#E74C3C", "#2ECC71"), lwd = 2, lty = 2)
  legend("topright", legend = c("Mean", "Median"),
         col = c("#E74C3C", "#2ECC71"), lty = 2, lwd = 2, bty = "n")
  hist(log10(x), breaks = 40, col = "#9B59B6", border = "white",
       main = "Histogram (log10 scale)", xlab = paste("log10 of", num_vars[[v]]))
  boxplot(x, log = "y", col = "#2ECC71", horizontal = FALSE,
          main = "Boxplot (log scale)", ylab = num_vars[[v]])
  par(mfrow = c(1, 1))
}
```


## 4.3 Interpretation by Variable

**Planet radius (`pl_rade`, n = 4,741).**
*Center:* The median radius is 2.47 Earth radii, while the mean is 4.50. The mean is almost twice the median because a smaller group of very large planets pulls it up, so the median describes a typical planet better.
*Variability:* The middle 50% of planets lie between 1.64 and 4.19 Earth radii (IQR = 2.55). The SD (4.94) show high relative variability, and the range (86.90) is dominated by a few extreme values.
*Shape:* The distribution is right-skewed. On the log10 scale it is bimodal: a main peak at about 2-3 Earth radii and a second peak near 12 Earth radii, close to Jupiter's size. This suggests two populations (small planets and gas giants), so neither the mean nor the median describes the second group.
*Outliers:* The boxplot rule flags 906 planets (19.1%), all above the upper fence of about 8.0 Earth radii; there are no lower outliers. This share is large because the gas-giant group lies above the fence. These planets are a real population and are retained. The maximum of 87.21 Earth radii (about 7.8 Jupiter radii) is far larger than any normal planet.

**Planet mass (`pl_bmasse`, n = 3,158).**
*Center:* The median is 171.2 Earth masses and the mean is 766.4, so the mean is about 4.5 times the median.
*Variability:* The IQR is 683.9 (Q1 = 12.2, Q3 = 696.0), the SD is 1,495, the highest relative variability among the radius, mass and temperature variables. The values range from 0.02 to 9,534.9 Earth masses.
*Shape:* The distribution is right-skewed. On the raw scale most planets fall in the first bar, and on the log10 scale the distribution is bimodal, with peaks at about 8 Earth masses and about 300 Earth masses (roughly one Jupiter mass). This agrees with the two populations seen in the radius.
*Outliers:* There are 422 outliers (13.4%) above the fence of about 1,722 Earth masses, with none below. The largest, about 9,535 Earth masses (≈ 30 Jupiter masses), is near the planet/brown-dwarf border. They are genuine massive objects and are retained, but they strongly affect the mean and SD. Some masses are lower bounds, which should be considered

**Orbital period (`pl_orbper`, n = 5,994).**
*Center:* The median period is 10.78 days, while the mean is 4,337.6 days. The mean is about 400 times larger than the median, so it is meaningless as a typical value.
*Variability:* The middle 50% lie between 4.31 and 38.55 days (IQR = 34.23). In contrast, the SD is 160,700 days, because the maximum is 8,040,000 days (about 22,000 years) and the minimum is only 0.09 days (about 2 hours).
*Shape:* The raw histogram is a single bar, so it is not informative. On the log10 scale the distribution has one peak at about 5-10 days and a long right tail, so it remains right-skewed even after the transformation.
*Outliers:* There are 1,041 outliers (17.4%) above the upper fence of about 90 days, and no lower outliers. A few values above 10⁵ days are extremely long and uncertain, probably from directly imaged planets on wide orbits. They are retained but flagged, and the median and IQR are used instead of the mean and SD.

**Semi-major axis (`pl_orbsmax`, n = 3,926).**
*Center:* The median is 0.12 AU and the mean is 20.11 AU, so a typical planet orbits much closer to its star than Mercury does (0.39 AU).
*Variability:* The IQR is 0.78 AU (Q1 = 0.05, Q3 = 0.84), but the SD is 405 AU, driven by a maximum of 19,000 AU.
*Shape:* The skewness is 35.73. The raw histogram is a single bar. On the log10 scale there is a large peak at about 0.05 AU and a smaller second bump at about 2 AU, and a long right tail.
*Outliers:* There are 606 outliers (15.4%) above the fence of about 2.0 AU. These include wide-orbit planets up to 19,000 AU. They are retained, but very large values are likely uncertain. (The minimum prints as 0.00 only because of rounding.)

**Equilibrium temperature (`pl_eqt`, n = 1,753).**
*Center:* The median is 900 K and the mean is 1,001 K, so the two are fairly close and the typical planet is hot.
*Variability:* Variability is moderate: IQR = 800 K (Q1 = 573, Q3 = 1,373), SD = 540 K much lower than for the other variables. The values range from 50 K to 3,921 K.
*Shape:* The skewness (0.82) shows only a mild right skew, and the histogram has a single peak at about 400-600 K. Unlike the other variables, the log transformation is not helpful here, because the log10 histogram becomes left-skewed, so the original scale is appropriate.
*Outliers:* The upper fence is about 2,573 K, so planets above that are flagged (the lower fence is below 0, so there are none on the low side). These are ultra-hot planets, are physically plausible, and are retained.

**Overall.** Radius, mass, period and semi-major axis are right-skewed, with the mean far above the median, so the median and IQR are the appropriate summaries. They span several orders of magnitude and need log scales. Equilibrium temperature is much more symmetric and less variable. The outlier shares are high (13-19%). The sample sizes range from 1,753 to 5,994, so each summary describes the planets for which that property could be measured, not all planets.

# 5. Numerical Variables by Categories

## 5.1 Planet Radius (pl_rade) by Discovery Method (discoverymethod)
 - **Differences in Location**: The median planet radius differs across the major discovery-method groups. The Transit group has a lower median radius than some of the other discovery methods. The median is useful for this comparison because planet radius has a skewed distribution and can contain extreme values.
 - **Differences in Variability and Shape**:The distributions also differ in their spread and shape. The Transit group has many observations concentrated at smaller planet radii, with a longer tail toward larger radii. The boxplot also shows differences in the IQR and the presence of potential outliers across discovery methods.
 To make the comparison readable, the four most frequent discovery methods were retained as separate categories and all remaining methods were combined into an "Other" category. This reduces the number of groups shown in the graph, but the "Other" category contains heterogeneous discovery methods and therefore should not be interpreted as a single detection technique.

```{r}
exo_clean <- exo %>% filter(default_flag == 1)

# ----------------------------------------------------------------------
#  Log Planet Radius (pl_rade) across Discovery Method (discoverymethod)
# ----------------------------------------------------------------------
# Categorical grouping: Top 4 most frequent discovery methods vs Other
top_methods <- exo_clean %>%
  count(discoverymethod) %>%
  top_n(4, n) %>%
  pull(discoverymethod)

exo_clean <- exo_clean %>%
  mutate(disc_method_cat = ifelse(discoverymethod %in% top_methods, discoverymethod, "Other"))

# Grouped Summary Table: Location and Spread
radius_by_method <- exo_clean %>%
  filter(!is.na(pl_rade)) %>%
  group_by(disc_method_cat) %>%
  summarise(
    Count = n(),
    Mean = mean(pl_rade, na.rm = TRUE),
    Median = median(pl_rade, na.rm = TRUE),
    SD = sd(pl_rade, na.rm = TRUE),
    IQR = IQR(pl_rade, na.rm = TRUE),
    Min = min(pl_rade, na.rm = TRUE),
    Max = max(pl_rade, na.rm = TRUE)
  )
radius_by_method


# Side-by-side Boxplot (Log10 scale due to skewness)
ggplot(exo_clean %>% filter(!is.na(pl_rade)), 
       aes(x = disc_method_cat, y = pl_rade, fill = disc_method_cat)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.3) +
  scale_y_log10() +
  labs(title = "Planet Radius Distribution by Discovery Method",
       x = "Discovery Method",
       y = "Planet Radius (Earth Radii, Log Scale)",
       fill = "Method") +
  theme_minimal() +
  theme(legend.position = "none")

Q1 <- quantile(exo_clean$pl_rade, 0.25, na.rm = TRUE)
Q3 <- quantile(exo_clean$pl_rade, 0.75, na.rm = TRUE)
IQR_value <- Q3 - Q1

outliers <- exo_clean %>%
  filter(
    pl_rade < Q1 - 1.5 * IQR_value |
    pl_rade > Q3 + 1.5 * IQR_value
  )

outliers %>%
  select(pl_name, pl_rade, discoverymethod)
```

## 5.2 Orbital Period (pl_orbper) across Stellar Multiplicity (sy_snum)
 - The distribution of orbital period was compared across planetary systems with different numbers of stars. The sy_snum variable was grouped into three categories: Single Star (sy_snum = 1), Binary Star System (sy_snum = 2), and Multi-Star (3+) (sy_snum ≥ 3). Only observations with available orbital-period values were included in this comparison.
 - **Differences in Location**: The median orbital period is 10.6 days for Single Star systems, 12.8 days for Binary Star Systems, and 13.7 days for Multi-Star systems. Therefore, the typical orbital period is somewhat longer in systems with two or more stars.
The means are much larger than the medians in all three groups. For example, the mean for Binary Star Systems is 16,857 days, compared with a median of only 12.8 days. This large difference suggests that a relatively small number of planets with very long orbital periods strongly affect the mean. Therefore, the median is more useful than the mean for comparing the typical orbital period across these groups.
 - **Differences in Variability and Shape**:Differences in Variability and Shape**
There are substantial differences in variability between the groups. The IQR is 30.8 days for Single Star systems, 213 days for Binary Star Systems, and 252 days for Multi-Star systems. This shows that the middle 50% of orbital periods are much more spread out in the Binary and Multi-Star groups.
The standard deviations are also very large, especially for Binary Star Systems. The standard deviation is 147,179 days for Single Star systems, 278,352 days for Binary Star systems, and 12,515 days for Multi-Star systems. These large values indicate that extreme orbital-period observations have a strong effect on the standard deviation.
 Overall, Most discovery-method groups have right-skewed distributions, as shown by the large differences between the means and medians. The boxplot and density plot on the logarithmic scale make differences in spread and the presence of extreme values easier to see. These results show an association between stellar multiplicity and the distribution of orbital periods, but they do not establish that the number of stars causes differences in orbital period.
 
```{r}
exo_clean <- exo_clean %>%
  mutate(snum_cat = factor(case_when(
    sy_snum == 1 ~ "Single Star",
    sy_snum == 2 ~ "Binary Star System",
    sy_snum >= 3 ~ "Multi-Star (3+)",
    TRUE ~ NA_character_
  )))

period_by_snum <- exo_clean %>%
  filter(!is.na(pl_orbper) & !is.na(snum_cat)) %>%
  group_by(snum_cat) %>%
  summarise(
    Count = n(),
    Mean = mean(pl_orbper, na.rm = TRUE),
    Median = median(pl_orbper, na.rm = TRUE),
    SD = sd(pl_orbper, na.rm = TRUE),
    IQR = IQR(pl_orbper, na.rm = TRUE)
  )
print(period_by_snum)

# Grouped Density Plot
ggplot(exo_clean %>% filter(!is.na(pl_orbper) & !is.na(snum_cat)), 
       aes(x = pl_orbper, fill = snum_cat)) +
  geom_density(alpha = 0.4) +
  scale_x_log10() +
  labs(title = "Density Distribution of Orbital Period by Stellar Multiplicity",
       x = "Orbital Period (Days, Log Scale)",
       y = "Density",
       fill = "Stellar System") +
  theme_minimal()

ggplot(
  exo_clean %>% filter(!is.na(pl_orbper), !is.na(snum_cat)),
  aes(x = snum_cat, y = pl_orbper)
) +
  geom_boxplot() +
  scale_y_log10() +
  labs(
    title = "Orbital Period by Stellar Multiplicity",
    x = "Stellar System",
    y = "Orbital Period (Days, Log Scale)"
  ) +
  theme_minimal()

```

# 6: Relationships Between Numerical Variables
## 6.1 Distance from Star (pl_orbsmax) vs. Orbital Period (pl_orbper)
- **Direction and Strength**: There is a strong positive relationship between a planet's semi-major axis and its orbital period. In general, planets with larger semi-major axes tend to have longer orbital periods. The Pearson correlation coefficient indicates a strong positive association between the two variables.
- **Form and Outliers**: On the logarithmic scales, the scatterplot shows a clear positive and approximately linear pattern.This indicates that the relationship is well described by a power-law-type pattern on the log-log scale. A small number of observations appear far from the main concentration of points, particularly at very large semi-major axes.

```{r}
# Semi-Major Axis vs. Orbital Period

# Filter out missing values, i.e complete-case analysis for this pair.
pair1 <- exo_clean %>% 
  filter(!is.na(pl_orbsmax) & !is.na(pl_orbper))

# Calculate Pearson Correlation
cor_pair1 <- cor(pair1$pl_orbsmax, pair1$pl_orbper, method = "pearson")
cat("Correlation between Distance and Period:", round(cor_pair1, 2), "\n")

# Simple Scatterplot
ggplot(pair1, aes(x = pl_orbsmax, y = pl_orbper)) +
  geom_point(alpha = 0.5, color = "steelblue") +
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  labs(
    title = "Orbital Period vs. Distance from Host Star",
    x = "Distance from Star (AU, Log Scale)",
    y = "Orbital Period (Days, Log Scale)"
  ) +
  theme_minimal()
```

## 6.2 Planet Mass (pl_bmasse) vs. Planet Radius (pl_rade)
- **Direction and Strength**: There is a positive relationship between planet mass and planet radius. In general, planets with larger masses tend to have larger radii, although the points show considerable variation around the overall trend.
- **Form and Outliers**: On the log-log scale, the scatterplot shows an overall upward trend, but the relationship is not perfectly linear. The spread of the observations changes across the range of mass and radius, and several observations appear separated from the main pattern. These observations can be considered potential outliers and should be investigated rather than automatically removed.
```{r}
# Filter out missing values
pair2 <- exo_clean %>% 
  filter(!is.na(pl_bmasse) & !is.na(pl_rade))

# Calculate Pearson Correlation
cor_pair2 <- cor(pair2$pl_bmasse, pair2$pl_rade, method = "pearson")
cat("Correlation between Mass and Radius:", round(cor_pair2, 2), "\n")

# Simple Scatterplot
ggplot(pair2, aes(x = pl_bmasse, y = pl_rade)) +
  geom_point(alpha = 0.5, color = "darkgreen") +
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(method = "lm", color = "orange", se = FALSE) +
  labs(
    title = "Planet Mass vs. Planet Radius",
    x = "Planet Mass (Earth Mass, Log Scale)",
    y = "Planet Radius (Earth Radii, Log Scale)"
  ) +
  theme_minimal()
```


## 6.3 Planet Mass vs. Planet Radius by Discovery Method

To investigate whether the discovery method changes the relationship between planet mass and radius, the relationship from Section 6.2 was compared across the main discovery-method groups.

The scatterplot shows a positive relationship between mass and radius across the discovery methods. However, the groups are distributed differently across the plot, with some methods concentrated in particular ranges of mass and radius. The fitted lines allow the overall relationship to be compared between the groups.

Overall, the positive relationship between mass and radius remains visible across the discovery methods, but the distribution and spread of the observations differ between groups. This suggests that discovery method is associated with differences in the observed mass-radius relationship, although this analysis does not establish a causal effect.

```{r}
## 6.3 Planet Mass vs. Planet Radius by Discovery Method

pair3 <- exo_clean %>%
  filter(
    !is.na(pl_bmasse),
    !is.na(pl_rade),
    !is.na(disc_method_cat)
  )

ggplot(pair3,
       aes(x = pl_bmasse,
           y = pl_rade,
           color = disc_method_cat)) +
  geom_point(alpha = 0.5) +
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Planet Mass vs. Radius by Discovery Method",
    x = "Planet Mass (Earth Mass, Log Scale)",
    y = "Planet Radius (Earth Radii, Log Scale)",
    color = "Discovery Method"
  ) +
  theme_minimal()
```





# 7. Distributional Analysis and QQ Plots

We study **planet equilibrium temperature (`pl_eqt`)**, measured in Kelvin.

Our questions are:

1. Can equilibrium temperature be reasonably described by a Normal distribution?
2. Does its distribution change by discovery method?

## 7.1 Overall Distribution

First, we remove missing values.

```{r point7-data}
eqt <- exo_clean$pl_eqt[!is.na(exo_clean$pl_eqt)]

length(eqt)
```

We calculate the main descriptive statistics.

```{r point7-statistics}
eqt_mean <- mean(eqt)
eqt_median <- median(eqt)
eqt_min <- min(eqt)
eqt_max <- max(eqt)
eqt_range <- eqt_max - eqt_min
eqt_q1 <- quantile(eqt, 0.25)
eqt_q3 <- quantile(eqt, 0.75)
eqt_iqr <- IQR(eqt)
eqt_sd <- sd(eqt)



cat("Number of observations:", length(eqt), "\n")
cat("Mean:", round(eqt_mean, 2), "K\n")
cat("Median:", round(eqt_median, 2), "K\n")
cat("Minimum:", eqt_min, "K\n")
cat("Maximum:", eqt_max, "K\n")
cat("Range:", round(eqt_range, 2), "K\n")
cat("Q1:", round(eqt_q1, 2), "K\n")
cat("Q3:", round(eqt_q3, 2), "K\n")
cat("IQR:", round(eqt_iqr, 2), "K\n")
cat("Standard deviation:", round(eqt_sd, 2), "K\n")
```

The mean is higher than the median and the skewness is positive, suggesting a right-skewed distribution.

### Histogram

```{r point7-histogram, fig.width=9, fig.height=5}
hist(
  eqt,
  breaks = 40,
  col = "#5B8FF9",
  border = "white",
  main = "Distribution of Planet Equilibrium Temperature",
  xlab = "Equilibrium Temperature (K)",
  ylab = "Number of Planets"
)

abline(
  v = eqt_mean,
  col = "#E74C3C",
  lwd = 3,
  lty = 2
)

abline(
  v = eqt_median,
  col = "#2ECC71",
  lwd = 3,
  lty = 2
)

legend(
  "topright",
  legend = c(
    paste("Mean =", round(eqt_mean, 1), "K"),
    paste("Median =", round(eqt_median, 1), "K")
  ),
  col = c("#E74C3C", "#2ECC71"),
  lty = 2,
  lwd = 3,
  bty = "n"
)
```

The histogram has a long upper tail, so the distribution is strongly right-skewed.

## 7.2 Normal Distribution and QQ Plot

We use the Normal distribution as a theoretical reference to see whether it is a reasonable description of the data.

```{r point7-overall-qq, fig.width=8, fig.height=6}
qqnorm(
  eqt,
  main = "Normal QQ Plot of Planet Equilibrium Temperature",
  xlab = "Theoretical Normal Quantiles",
  ylab = "Observed Quantiles",
  pch = 16,
  col = "#5B8FF9"
)

qqline(
  eqt,
  col = "#E74C3C",
  lwd = 3
)
```

The points are closer to the line in the middle but deviate systematically in the tails, especially the upper tail.

Therefore, equilibrium temperature is **not reasonably described by a Normal distribution**.

The extreme upper observations are potentially unusual, but we do not automatically consider them errors.

## 7.3 Distribution by Discovery Method

We compare:

- Transit
- Radial Velocity
- Imaging

```{r point7-groups}
transit <- exo_clean$pl_eqt[
  exo_clean$discoverymethod == "Transit" &
    !is.na(exo_clean$pl_eqt)
]

rv <- exo_clean$pl_eqt[
  exo_clean$discoverymethod == "Radial Velocity" &
    !is.na(exo_clean$pl_eqt)
]

imaging <- exo_clean$pl_eqt[
  exo_clean$discoverymethod == "Imaging" &
    !is.na(exo_clean$pl_eqt)
]

cat("Transit:", length(transit), "\n")
cat("Radial Velocity:", length(rv), "\n")
cat("Imaging:", length(imaging), "\n")
```

Transit has more than 17,000 observations, while Imaging has about 101, so the Imaging results should be interpreted more carefully.

### Boxplot

```{r point7-boxplot, fig.width=9, fig.height=6}
boxplot(
  transit,
  rv,
  imaging,
  names = c("Transit", "Radial Velocity", "Imaging"),
  col = c("#5B8FF9", "#9B59B6", "#2ECC71"),
  border = "black",
  main = "Equilibrium Temperature by Discovery Method",
  xlab = "Discovery Method",
  ylab = "Equilibrium Temperature (K)",
  las = 1
)
```

Imaging has the highest center and the largest spread.

Transit and Radial Velocity have lower centers and more right-skewed distributions.

### Numerical Summary

```{r point7-group-statistics}
transit_mean <- mean(transit)
transit_median <- median(transit)
transit_sd <- sd(transit)
transit_iqr <- IQR(transit)



rv_mean <- mean(rv)
rv_median <- median(rv)
rv_sd <- sd(rv)
rv_iqr <- IQR(rv)



imaging_mean <- mean(imaging)
imaging_median <- median(imaging)
imaging_sd <- sd(imaging)
imaging_iqr <- IQR(imaging)



group_summary <- data.frame(
  Method = c(
    "Transit",
    "Radial Velocity",
    "Imaging"
  ),
  Count = c(
    length(transit),
    length(rv),
    length(imaging)
  ),
  Mean = c(
    transit_mean,
    rv_mean,
    imaging_mean
  ),
  Median = c(
    transit_median,
    rv_median,
    imaging_median
  ),
  SD = c(
    transit_sd,
    rv_sd,
    imaging_sd
  ),
  IQR = c(
    transit_iqr,
    rv_iqr,
    imaging_iqr
  )
)

knitr::kable(
  group_summary,
  digits = 2,
  caption = "Equilibrium Temperature by Discovery Method"
)
```

Imaging has the highest median and largest IQR.

Transit and Radial Velocity have positive skewness, while Imaging has skewness close to zero.

## 7.4 QQ Plots by Discovery Method

```{r point7-group-qq, fig.width=12, fig.height=5}
par(mfrow = c(1, 3))

qqnorm(
  transit,
  main = "Transit",
  xlab = "Theoretical Normal Quantiles",
  ylab = "Observed Quantiles",
  pch = 16,
  col = "#5B8FF9"
)

qqline(
  transit,
  col = "#E74C3C",
  lwd = 2
)

qqnorm(
  rv,
  main = "Radial Velocity",
  xlab = "Theoretical Normal Quantiles",
  ylab = "Observed Quantiles",
  pch = 16,
  col = "#9B59B6"
)

qqline(
  rv,
  col = "#E74C3C",
  lwd = 2
)

qqnorm(
  imaging,
  main = "Imaging",
  xlab = "Theoretical Normal Quantiles",
  ylab = "Observed Quantiles",
  pch = 16,
  col = "#2ECC71"
)

qqline(
  imaging,
  col = "#E74C3C",
  lwd = 2
)

par(mfrow = c(1, 1))
```

Transit and Radial Velocity show stronger curvature and right-skewness.

Imaging is appears more symmetric and closer to a linear pattern in the middle, but its sample size is much smaller.

## 7.5 Final Conclusion

The three discovery methods have clearly different equilibrium-temperature distributions.

- **Location:** Imaging has the highest median.
- **Spread:** Imaging has the largest IQR.
- **Skewness:** Transit and Radial Velocity are right-skewed.
- **Tails:** Transit has a particularly long upper tail.
- **Shape:** Imaging is more symmetric.

Overall, the differences are substantial rather than minor.

The results show an association between discovery method and equilibrium temperature, but they do not show that discovery method causes the differences.




#Updated

## 8. Integrated Exploratory Data Analysis

The previous sections looked at one or two variables at a time. Here we combine several variables to answer three questions that came out of the earlier results:

- the planet radius distribution is bimodal (small planets and gas giants);
- the discovery-method mix changes strongly over time (Section 3.5);
- orbital distance and period are strongly related (Section 6.1).

```{r s8-setup}
library(dplyr)
library(ggplot2)
library(scales)

# Rebuild helper variables so that this section does not depend on earlier edits
exo8 <- exo_clean %>%
  mutate(
    discoverymethod = as.character(discoverymethod),
    method_grp = ifelse(discoverymethod %in% c("Transit", "Radial Velocity",
                                               "Microlensing", "Imaging"),
                        discoverymethod, "Other"),
    disc_era = cut(disc_year,
                   breaks = c(-Inf, 2009, 2014, 2019, Inf),
                   labels = c("Before 2010", "2010-2014", "2015-2019", "2020 onward")),
    # Planet size classes (Earth radii)
    size_class = cut(pl_rade,
                     breaks = c(0, 2, 6, Inf),
                     labels = c("Small (<2)", "Mid-size (2-6)", "Giant (>=6)"),
                     right = FALSE)
  )
```

### 8.1 Question 1: Has the typical size of newly discovered planets changed over time, and is this a real change or a detection effect?

**Variables:** planet radius (`pl_rade`), discovery period (`disc_era`), size class, and discovery method (`discoverymethod`) as a check.

**Methods:** grouped summaries by era, share of each size class by era, boxplots on the log scale, and a repeat of the analysis using only Transit discoveries to see whether a change in method explains the pattern.

```{r s8-q1-table}
q1 <- exo8 %>% filter(!is.na(pl_rade), !is.na(disc_era))

q1_summary <- q1 %>%
  group_by(disc_era) %>%
  summarise(
    n         = n(),
    Median    = median(pl_rade),
    Q1        = quantile(pl_rade, 0.25),
    Q3        = quantile(pl_rade, 0.75),
    IQR       = IQR(pl_rade),
    Pct_Small = 100 * mean(size_class == "Small (<2)"),
    Pct_Mid   = 100 * mean(size_class == "Mid-size (2-6)"),
    Pct_Giant = 100 * mean(size_class == "Giant (>=6)")
  )
knitr::kable(q1_summary, digits = 2,
             caption = "Planet radius (Earth radii) by discovery period")
```

```{r s8-q1-plots, fig.width=9, fig.height=4.5}
# (a) Boxplot of radius by era (log scale)
ggplot(q1, aes(x = disc_era, y = pl_rade, fill = disc_era)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.3, show.legend = FALSE) +
  scale_y_log10() +
  labs(title = "Planet Radius by Discovery Period",
       x = "Discovery Period", y = "Planet Radius (Earth radii, log scale)") +
  theme_minimal()

# (b) Size-class mix by era
ggplot(q1, aes(x = disc_era, fill = size_class)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = percent) +
  labs(title = "Size-Class Mix of Discovered Planets by Period",
       x = "Discovery Period", y = "Proportion of Planets", fill = "Size class") +
  theme_minimal()
```

```{r s8-q1-transit}
# Does the pattern survive when only one method (Transit) is used?
q1_transit <- q1 %>%
  filter(discoverymethod == "Transit") %>%
  group_by(disc_era) %>%
  summarise(n = n(),
            Median = median(pl_rade),
            Pct_Giant = 100 * mean(size_class == "Giant (>=6)"))
knitr::kable(q1_transit, digits = 2,
             caption = "Transit discoveries only: radius by period")

# Radial-velocity masses over time (sensitivity of the method)
exo8 %>%
  filter(discoverymethod == "Radial Velocity", !is.na(pl_bmasse), !is.na(disc_era)) %>%
  group_by(disc_era) %>%
  summarise(n = n(), Median_mass = median(pl_bmasse)) %>%
  knitr::kable(digits = 1,
               caption = "Radial Velocity discoveries: median planet mass (Earth masses)")
```

**Results and interpretation.** The size of discovered planets changed dramatically. Before 2010, the median radius was about 13.7 Earth radii and 93.2% of planets with a measured radius were giants (n = 73). From 2010 to 2014 the median fell to about 2.35 Earth radii, and in 2015-2019 to about 2.26, with 38-42% of planets in the small class and only 16-18.5% giants. In 2020 onward the median rose slightly to about 2.76 and the giant share rose to 26.8%, so the trend is not monotone. The spread also changed: the IQR interval was [12.3,15.6] before 2010  but in 2020 onward was [7.25,1.92], so the discovery of giants was mostly before . The same pattern appears among Transit discoveries alone (giant share 94.7% → 17.6% → 15.9% → 26.5%), so the change cannot be explained only by a shift in discovery method. The median mass of Radial Velocity planets also fell steadily across the four periods (about 526 → 350 → 213 → 75 Earth masses), which shows the same effect for a second method. Radial velocity detects a planet by the tiny wobble its gravity causes in the star. The size of the wobble depends on the planet's mass, so mass comes out of the measurement directly.

**What this tells us:** early surveys could only detect large, massive planets, and as instruments and surveys improved, much smaller planets became detectable. **What it does not tell us:** that the *population* of planets changed. The pre-2010 group is also small (n = 73 with radius), so its summaries are less stable, and the small rise after 2020 may reflect specific surveys targeting giant planets.


### 8.2 Question 2: Does the relationship between equilibrium temperature and orbital distance depend on the temperature of the host star?

**Variables:** equilibrium temperature (`pl_eqt`), semi-major axis (`pl_orbsmax`), and host star effective temperature (`st_teff`), split into three equal-sized groups (Cool, Medium, Hot).

**Methods:** log-log scatterplot coloured by star class, group summaries, correlations and fitted slopes within each class.

```{r s8-q3-data}
q3 <- exo8 %>%
  filter(!is.na(pl_eqt), !is.na(pl_orbsmax), !is.na(st_teff))

# Three equal-sized groups of host stars by effective temperature
cuts <- quantile(q3$st_teff, probs = c(0, 1/3, 2/3, 1))
q3 <- q3 %>%
  mutate(star_class = cut(st_teff, breaks = cuts, include.lowest = TRUE,
                          labels = c("Cool host", "Medium host", "Hot host")))
# counting the rows of q3 which includes the data where all key attributes for our analysis are present
nrow(q3)

q3 %>%
  group_by(star_class) %>%
  summarise(n = n(),
            Teff_min = min(st_teff), Teff_max = max(st_teff),
            Median_eqt = median(pl_eqt),
            Median_smax = median(pl_orbsmax)) %>%
  knitr::kable(digits = 3, caption = "Host-star classes and planet summaries")
```

```{r s8-q3-plot, fig.width=8, fig.height=5}
ggplot(q3, aes(x = pl_orbsmax, y = pl_eqt, color = star_class)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_x_log10() +
  scale_y_log10() +
  labs(title = "Equilibrium Temperature vs. Orbital Distance by Host-Star Class",
       x = "Semi-Major Axis (AU, log scale)",
       y = "Equilibrium Temperature (K, log scale)",
       color = "Host star") +
  theme_minimal()
```

```{r s8-q3-cor}
cat("Overall correlation (log-log) =", round(cor(log10(q3$pl_orbsmax), log10(q3$pl_eqt)), 2), "\n")

q3 %>%
  group_by(star_class) %>%
  summarise(n = n(),
            r_log = cor(log10(pl_orbsmax), log10(pl_eqt))) %>%
  knitr::kable(digits = 2, caption = "Log-log correlation and slope by host-star class")
```

**Results and interpretation.** Equilibrium temperature tends to decrease with distance, but the strength depends on the host. Overall (n = 1,542) the log-log correlation is only -0.34. Within host groups, the relationship is strong for medium-temperature hosts (r = -0.84) and for hot hosts (r = -0.62), but almost absent for cool hosts (r = -0.11). The classes also differ in level: the median planet temperature is about 583 K around cool hosts, 952 K around medium hosts and 1,431 K around hot hosts, even though the median orbital distance is similar (0.05-0.07 AU) in all three. So, at a given distance, planets around hotter stars are hotter, and this is the main reason why the overall correlation is weak. Physical models predict a slope of about -0.5 in this log-log relation, and only the medium group is relatively close to this.

**What this tells us:** the host star must be taken into account when relating planet temperature to orbital distance. **What it does not tell us:** `pl_eqt` is not measured directly. It is *calculated* from the star's temperature and radius and the orbital distance, so part of this relationship holds by construction. The sample is also limited to planets that have all three variables (1,542 of 6,366), mostly transiting planets close to their stars, so the cool-host result may reflect a narrow range of distances rather than a true absence of relationship.

## 10. Conclusions

In this project we analysed the NASA Exoplanet Archive, using one default parameter set per planet (6,366 planets). The main findings are below.

**1. Most important characteristics of the dataset.**
The data are very unbalanced. Transit discovers 74% of planets and Radial Velocity 19%, 91% of planets orbit single stars, and 94% were discovered after 2010. Many measurements are missing (for example, 81% of planet masses), so each analysis uses a different number of planets. The numerical variables (radius, mass, period, distance) are strongly right-skewed and cover several orders of magnitude, so the median and IQR describe them better than the mean and SD. On the log scale, radius and mass are bimodal, which suggests two populations: small planets and gas giants.

**2. Patterns and associations.**
The discovery method is associated with the stellar multiplicity (about 6% of Transit planets are in binary systems, against 16.5% for Imaging). The method mix also changed over time: Radial Velocity dominated before 2010, while Transit dominates afterwards. The size of the discovered planets changed as well: the median radius fell from about 13.7 Earth radii before 2010 to about 2.3-2.8 later. Giant planets are concentrated at short periods (about 76% of giants orbit in under 10 days).

**3. Differences between groups.**
Planet radius differs a lot between methods (median about 15 Earth radii for Imaging against 2.5 for Transit). Orbital period is longer and much more spread out in binary and multi-star systems (IQR about 213-252 days against 31 days for single stars). Equilibrium temperature also differs (median about 1,608 K for Imaging, 939 K for Transit and 420 K for Radial Velocity).

**4. Relationships between variables.**
Orbital distance and orbital period have a very strong relationship on the log scale (r = 0.995), and mass and radius are strongly related on the log scale (r = 0.86). A planet's equilibrium temperature decreases with distance, but how strongly depends on the host star (r = -0.84 for medium-temperature hosts and r = -0.11 for cool hosts).

**5. Distributions across categories.**
Yes, the distributions clearly change across categories. The equilibrium-temperature distributions differ in location, spread and skewness between the three methods, and the differences are substantial. Imaging is the highest and the most symmetric (but has few planets), while Transit and Radial Velocity are right-skewed.

**6. Theoretical distributions.**
Equilibrium temperature is not reasonably described by a Normal distribution. Its QQ plot deviates from the line, mainly in the upper tail, and the groups by method are even further from normal. The other variables are even more skewed, so none of them is close to Normal on the original scale.

**7. Surprising findings.**
Three results surprised us. First, the raw correlations can be very misleading: mass and radius have r = 0.36 on the raw scale but r = 0.86 on the log scale, because of extreme values. Second, pooled results can hide subgroups: radius and period are almost uncorrelated overall (r = 0.04), but they have opposite trends for small planets (+0.28) and giants (-0.34). Third, the share of giant planets went up again after 2020, so the change in planet size over time was not a simple decrease.

**8. Limitations.**
- The archive contains planets that we are *able to detect*, not a random sample of all planets, so many patterns (size over time, hot Jupiters, method differences) are mostly detection bias.
- There are a lot of missing values, so the analysis uses available cases only, and the planets with all variables may not represent the rest.
- Some groups are small (Imaging, Other, and Radial Velocity planets with radius), so their summaries are unstable.
- Some masses are only minimum masses (Msini), and the most extreme orbital values are uncertain.
- Equilibrium temperature is calculated from other variables, not measured directly, and the size classes (2 and 6 Earth radii) are arbitrary.
- We kept all outliers because they are real objects, but they strongly affect means and standard deviations.

Finally, this is an exploratory analysis. It describes patterns and associations, but it does not establish causation.





