---
title: "CS 108 Statistics — Project 1"
date: "2026-09-27"
output:
  html_document:
    toc: true
    toc_depth: 2
  
---

```
{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 8, fig.height = 5)
```

# 5. Numerical Variables by Categories

In this section, we investigate whether the distribution of planet radius differs across different discovery methods. The numerical variable is pl_rade, measured in Earth radii, and the categorical variable is discoverymethod.

## 5.1 Planet Radius by Discovery Method
We first calculate descriptive statistics for planet radius within each discovery method.
```{r load-data}
library(readr)
library(ggplot2)
library(dplyr)

exo <- read_csv("PS_2026.09.28_04.07.20.csv",, comment = "#",  show_col_types = FALSE)
```

```{r}

boxplot(pl_orbper ~ disc_locale,
        data = exo,
        main = "Orbital Period by Discovery Locale",
        xlab = "Discovery Locale",
        ylab = "Orbital Period (days)")
table(exo$discoverymethod, exo$disc_locale)
ggplot(exo, aes(x = discoverymethod, fill = disc_locale)) +
  geom_bar() +
  labs(
    title = "Discovery Method by Observation Location",
    x = "Discovery Method",
    y = "Number of Exoplanets",
    fill = "Discovery Location"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
```{r}
names(exo)

barplot(table(exo$discoverymethod, exo$disc_year), col=rainbow(10), beside = TRUE)

```


## 5.2 Orbital Period by Discovery Method
 We next investigate whether orbital periods differ across discovery
methods.


